---
title: "Malaysian Elections"
subtitle: "Part three: Online Sentiments"
author: "Sean Ng"
organization: "AIMdata"
date-modified: "13 May 2026"
execute: 
  echo: false
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 9)


library(tidyverse)
library(here)
library(janitor)
library(scales)
library(tidytext)
library(widyr)
library(ggraph)
library(patchwork)
library(kableExtra)
library(viridis)
library(stringr)
library(sf)
library(broom)
library(malaytextr)
library(topicmodels)

`%out%` <- Negate(`%in%`)
options(scipen = 100)
theme_set(theme_light())
range_wna <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

last_comma <- function(x, sep = ","){
  x <- sapply(strsplit(x, sep),"[")
  trimws(x[[length(x)]])
}
```

Najib last term 
5 May 2013 - 9 May 2018

Mahathir 
9 May 2018 - 28 February 2020

Muhyiddin
1 March 2020 - 20 August 2021

Ismail Sabri
21 August 2021 - 23 November 2022

Anwar Ibrahim 
24 November 2022 - present 



```{r}
threads <- read_csv("./data/malaysia_subreddit_threads.csv")

threads_2026 <- read_csv("./data/malaysia_subreddit_threads_2026.csv")

# Commented out because I've already taken a subset of it
# comments <- read_csv("./raw_data/malaysia_subreddit_comments.csv") |>
#    left_join(
#     threads |> 
#       distinct(thread_id, flair = link_flair_text, title), 
#     by = "thread_id"
#   ) |> 
#   rbind(read_csv("./data/malaysia_subreddit_comments_2026.csv") |> 
#           filter(created_utc > "2026-01-03 20:41:56 UTC") |> 
#           left_join(
#     threads_2026 |> 
#       distinct(thread_id, flair = link_flair_text, title), 
#     by = "thread_id"
#   )) |> 
#   mutate(flair = ifelse(is.na(flair) & str_detect(title, "daily random"),
#                                   "Daily Random", 
#                                   flair)) |> 
#   filter(
#     flair %in% c("Economy & Finance", "Politics", "Religion")
#   )
#
# comments |> write_csv("./data/malaysia_subreddit_comments_filtered.csv")

comments <- read_csv("./data/malaysia_subreddit_comments_filtered.csv")

malay_slang <- read_csv("./data/malay_slang.csv")

# malay slang dictionary compiled by https://data.mendeley.com/datasets/mgv2n2vcb9/3

malay_eng_73k <- read.table("https://dl.fbaipublicfiles.com/arrival/dictionaries/en-ms.txt", header = FALSE) |> 
  rename(english = V1, malay = V2)

```


Threads about religion are the most popular attract the most comments, giving additional credence to 
[Deivasagayam's](https://msocialsciences.com/index.php/mjssh/article/view/3248/2134) argument about the increasing importance of

convergence of race and religion. 

## Race and religion

```{r}
comments |>
  group_by(flair, thread_id) |>
  summarise(count = n_distinct(body), 
            .groups = "drop") |> 
  group_by(flair) |> 
  summarise(count = mean(count)) |>
  mutate(count = round(count, 2)) |> 
  ggplot(aes(x = count, y = flair)) + 
  geom_col(aes(fill = flair)) + 
  geom_text(aes(label = count), 
            hjust = "inward") + 
  scale_fill_viridis_d(guide = "none", 
                       option = "cividis") + 
  labs(x = "Mean number of comments per post", 
       y = "Post flair", 
       title = "Mean number of comments per post, by post flair")
  

```

<br>

```{r}
comments |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot")) |> 
  mutate(is_anwar = 
           ifelse(
             str_detect(title, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister") |
               str_detect(body, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister"), 
                           1, 0)) |> 
  mutate(count = 1) |> 
  group_by(is_anwar, flair) |> 
  summarise(count = sum(count)) |> 
  mutate(is_anwar = ifelse(is_anwar == 1, "Anwar related", "non-Anwar")) |> 
  ggplot(aes(x = count, y = fct_rev(is_anwar))) + 
  geom_col() + 
  geom_text(aes(label = comma(count)),
            hjust = "inward") + 
  facet_wrap(~ flair)
```

```{r}
anwar_words <- comments |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot")) |> 
  mutate(is_anwar = 
           ifelse(
             str_detect(title, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister") |
               str_detect(body, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister"), 
                           1, 0)) |> 
  filter(is_anwar == 1) |> 
  mutate(score_trend = case_when(
    score > 1 ~ "positive", 
    score == 1 ~ "neutral", 
    score <= 0 ~ "negative"
  )) |> 
  select(comment_id, thread_id, body, score, score_trend, flair) |> 
  mutate(body = remove_url(body)) |>
  unnest_tokens(word, body) |> 
  # Translations
  mutate(word = ifelse(word %in% malay_slang$slang, malay_slang$translation, word),
         word = ifelse(word %in% malay_eng_73k$malay, malay_eng_73k$english, word)) |> 
  mutate(word = str_replace(word, "kepada", "to")) |>
  filter(word %out% c("emote", "t5_2qh8b", "giphy", "it’s", "don’t")) |> 
  anti_join(stop_words, by = "word") |> 
  anti_join(malaystopwords, by = c("word" = "stopwords")) |> 
  add_count(word, flair) |> 
  filter(str_detect(word, "[a-z]")) 

israel_words <- comments |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot")) |> 
  mutate(is_israel = 
           ifelse(
             str_detect(title, "Israel|israel|Jew|jew|Palestin|palestin|Gaza|gaza") |
               str_detect(body, "Israel|israel|Jew|jew|Palestin|palestin|Gaza|gaza"), 
                           1, 0)) |> 
  mutate(score_trend = case_when(
    score > 1 ~ "positive", 
    score == 1 ~ "neutral", 
    score <= 0 ~ "negative"
  )) |> 
  filter(is_israel == 1) |> 
  select(comment_id, thread_id, body, score, score_trend, flair) |> 
  mutate(body = remove_url(body)) |>
  unnest_tokens(word, body) |> 
  # Translations
  mutate(word = ifelse(word %in% malay_slang$slang, malay_slang$translation, word),
         word = ifelse(word %in% malay_eng_73k$malay, malay_eng_73k$english, word)) |> 
  mutate(word = str_replace(word, "kepada", "to")) |>
  filter(word %out% c("emote", "t5_2qh8b", "giphy", "it’s", "don’t")) |> 
  anti_join(stop_words, by = "word") |> 
  anti_join(malaystopwords, by = c("word" = "stopwords")) |> 
  add_count(word, flair) |> 
  filter(str_detect(word, "[a-z]"))   
```

```{r}
comment_words <- comments |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot|this comment was removed due")) |> 
  mutate(is_anwar = 
           ifelse(
             str_detect(title, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister") |
               str_detect(body, "Anwar|anwar|pmx|PMX|Pmx|PM|pm|prime minister|Prime Minister"), 
                           1, 0)) |> 
  mutate(score_trend = case_when(
    score > 1 ~ "positive", 
    score == 1 ~ "neutral", 
    score <= 0 ~ "negative"
  )) |> 
  select(comment_id, thread_id, body, score, score_trend, flair) |> 
  mutate(body = remove_url(body)) |>
  unnest_tokens(word, body) |> 
  # Translations
  mutate(word = ifelse(word %in% malay_slang$slang, malay_slang$translation, word),
         word = ifelse(word %in% malay_eng_73k$malay, malay_eng_73k$english, word)) |> 
  mutate(word = str_replace(word, "kepada", "to")) |>
  filter(word %out% c("emote", "t5_2qh8b", "giphy", "it’s", "don’t")) |> 
  anti_join(stop_words, by = "word") |> 
  anti_join(malaystopwords, by = c("word" = "stopwords")) |> 
  add_count(word, flair) |> 
  filter(str_detect(word, "[a-z]")) 
```




```{r}

anwar_words |> 
  distinct(flair, word, n) |>
  filter(word %out% c("pm", "pmx", "it’s")) |> 
  group_by(flair) |> 
  slice_max(order_by = n, n = 15) |> 
  ggplot(aes(x = n, y = reorder_within(word, by = n, within = flair))) + 
  geom_col(aes(fill = flair)) + 
  facet_wrap(~ flair, scales = "free") +  
  scale_y_reordered() + 
  scale_fill_viridis_d(guide = "none", 
                       option = "cividis") + 
  labs(title = "Most common words associated with Anwar Ibrahim, by post flair", 
       x = "Number of occurences", 
       y = "", 
       caption = "Source: https://www.kaggle.com/datasets/bwandowando/reddit-rmalaysia-subreddit-dataset") +
  theme(strip.background = element_rect(fill = "grey20"))
```
<br>

Some of the most common terms that Anwar is mentioned alongside under the "Economics and Finance" flair include **subsidies**, **taxes**, **companies** and **prices**. Under "Politics", in addition to political coalitions like **PH** and **BN**, numerous *Reformasi* topics are also mentioned, including **money** and **corruption**. **Malay** and **Chinese** also indicate the predominance of racial politics; and **Rafizi** indicates talk about the latest of Anwar's intra-party struggles. 

Under "Religion", there is a wider spread of issues, 

<br>

## LDA 

```{r}
comments_bigrams <- comments |> 
  mutate(is_anwar = ifelse(str_detect(title, "Anwar|anwar|pmx|PMX|Pmx|PM|pm") |
                             str_detect(body, "Anwar|anwar|pmx|PMX|Pmx|PM|pm"), 
                           1, 0)) |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot|this comment was removed due")) |>
  mutate(body = str_replace_all(body, "non |Non ", "non-")) |> 
  mutate(score_trend = case_when(
    score > 1 ~ "positive", 
    score == 1 ~ "neutral", 
    score > 10 ~ "very positive", 
    score <= 0 ~ "negative"
  )) |> 
  select(comment_id, thread_id, body, score, score_trend, flair) |> 
  mutate(body = remove_url(body)) |> 
  unnest_tokens(bigram, body, token = "ngrams", n = 2) |> 
  add_count(bigram) |> 
  filter(!is.na(bigram)) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  mutate(word1 = ifelse(word1 %in% malay_slang$slang, malay_slang$translation, word1)) |>
  mutate(word2 = ifelse(word2 %in% malay_slang$slang, malay_slang$translation, word2)) |>
  filter(!word1 %in% stop_words$word & str_detect(word1, "[a-z]")) |> 
  filter(!word2 %in% stop_words$word & str_detect(word2, "[a-z]")) |> 
  count(word1, word2, thread_id, sort = TRUE) |> 
  unite(bigram, word1, word2, sep = " ") |> 
  bind_tf_idf(bigram, thread_id, n) |> 
  left_join(
    rbind(threads, threads_2026) |>
      distinct(thread_id, flair = link_flair_text, title) |>
      filter(flair %in% c("Politics", "Economy & Finance", "Religion")), 
    by = "thread_id"
  )
```




```{r}
comments_bigrams |> 
  filter(!str_detect(bigram, "gif|giphy|emote")) |> 
  mutate(bigram = str_trim(bigram)) |> 
  group_by(flair, bigram) |> 
  summarise(n = sum(n)) |> 
  slice_max(order_by = n, n = 20) |> 
  ungroup() |> 
  ggplot(aes(x = n, 
             y = tidytext::reorder_within(bigram, by = n, within = flair))) + 
  geom_col(aes(fill = flair)) + 
  facet_wrap(~ flair, scales = "free") +  
  scale_y_reordered() + 
  scale_fill_viridis_d(guide = "none", 
                       option = "cividis") + 
  labs(title = "Most common bigrams by post flair", 
       x = "Number of occurences", 
       y = "", 
       caption = "Source: https://www.kaggle.com/datasets/bwandowando/reddit-rmalaysia-subreddit-dataset") +
  theme(strip.background = element_rect(fill = "grey20"), 
        plot.caption = element_text(hjust = .5))
```

<br>

Subtstantial overlap between politics and religion, given the prominence of the term *ajaran sesat* (heresy)

We can also infer that there are a high proportion of non-Muslims (probably much higher than the national average) in the r/Malaysia subreddit, given the high frequency of mentions about Halal certification, a common talking point amongst Malaysian Chinese, many of whom tend to have strong feelings about Muslim dietary restrictions.  

Contentious issues like the middle east (genocide in Palestine, Iran war)

Common anti-Muslim talking points. 
Halal certification is 
Representative of a 

These issues 
Disastissfaction with Najib's house arrest


```{r}
comment_words |> glimpse()
  cast_dtm(thread_id, )

comments |> 
  filter(str_detect(body, "hang tuah|Hang Tuah")) |>
  sample_n(5) |> 
  pull(body)

comments_bigrams
```

## Network graphs


```{r}
set.seed(333)

anwar_words_network <- anwar_words |> 
  filter(n > 49) |> 
  pairwise_cor(word, comment_id, sort = TRUE) |> 
  filter(correlation >= .18) |> 
  left_join(
    anwar_words |> 
      filter(n > 49) |> 
      pairwise_count(word, comment_id, sort = TRUE), 
    by = c("item1", "item2")  
  ) |> 
  igraph::graph_from_data_frame() |> 
  ggraph(layout = "kk") +
  geom_edge_link(aes(alpha = correlation, edge_width = n), colour = "#457b9d", check_overlap = TRUE) +
  scale_edge_width_continuous(range = c(.1, 3), trans = "log10") +
  scale_alpha_continuous(range = c(0.01, 0.08)) +
  geom_node_point(colour = "#457b9d", alpha = 0.2, size = .5) +
  geom_node_text(aes(label = name), size = 2.7, alpha = .5, 
                 colour = "#caf0f8") +
  theme(legend.position = "none", 
        plot.caption = element_text(hjust = .5, colour = "#219ebc"), 
        panel.background = element_rect(fill = "#03071e"), 
        plot.background = element_rect(fill = "#1b263b"), 
        plot.title = element_text(colour = "#219ebc"), 
        plot.subtitle = element_text(colour = "#219ebc", size = 10)) + 
  labs(title = paste0("Network graph of comments related to Anwar Ibrahim, ", 
                      comments %>% {min(.$created_utc)} %>% 
                        round_date(unit = "day") %>% str_remove_all(., " UTC"), 
                      " to ", 
                       comments %>% {max(.$created_utc)} %>% 
                        round_date(unit = "day") %>% str_remove_all(., " UTC")), 
       subtitle = "Line thickness indicates number of comments using those words, transparency indicates correlation between words", 
       caption = "Source: https://www.kaggle.com/datasets/bwandowando/reddit-rmalaysia-subreddit-dataset")

ggsave(here("plots", "anwar_network_graph.png"), 
       width = 10, height = 6.5, units = "in", dpi = 300)

```

We can see from the network graph that religion is involved with Anwar, even if he tries to distance himself from it. 

```{r}
comments_bigrams |> 
  filter(!str_detect(bigram, "gif|giphy|emote")) |> 
  filter(n > 15) |> 
  pairwise_cor(bigram, thread_id, sort = TRUE) |> 
  filter(correlation >= .13) |> 
  left_join(
    comments_bigrams |> 
      filter(n > 15) |> 
      pairwise_count(bigram, thread_id, sort = TRUE), 
    by = c("item1", "item2")  
  ) |> 
  igraph::graph_from_data_frame() |> 
  ggraph(layout = "kk") +
  geom_edge_link(aes(alpha = correlation, edge_width = n), colour = "#457b9d", check_overlap = TRUE) +
  scale_edge_width_continuous(range = c(.1, 3), trans = "log10") +
  scale_alpha_continuous(range = c(0.01, 0.08)) +
  geom_node_point(colour = "#457b9d", alpha = 0.2, size = .5) +
  geom_node_text(aes(label = name), size = 2.7, alpha = .5, 
                 colour = "#caf0f8") +
  theme(legend.position = "none", 
        plot.caption = element_text(hjust = .5, colour = "#219ebc"), 
        panel.background = element_rect(fill = "#03071e"), 
        plot.background = element_rect(fill = "#1b263b"), 
        plot.title = element_text(colour = "#219ebc"), 
        plot.subtitle = element_text(colour = "#219ebc", size = 10)) + 
  labs(title = paste0("Network graph of comments related to Anwar Ibrahim, ", 
                      comments %>% {min(.$created_utc)} %>% 
                        round_date(unit = "day") %>% str_remove_all(., " UTC"), 
                      " to ", 
                       comments %>% {max(.$created_utc)} %>% 
                        round_date(unit = "day") %>% str_remove_all(., " UTC")), 
       subtitle = "Line thickness indicates number of comments using those words, transparency indicates correlation between words", 
       caption = "Source: https://www.kaggle.com/datasets/bwandowando/reddit-rmalaysia-subreddit-dataset")

ggsave(here("plots", "bigrams_network_graph.png"), 
       width = 10, height = 6.5, units = "in", dpi = 300)
```



```{r}
anwar_bigrams <- comments |> 
  mutate(is_anwar = ifelse(str_detect(title, "Anwar|anwar|pmx|PMX|Pmx|PM|pm") |
                             str_detect(body, "Anwar|anwar|pmx|PMX|Pmx|PM|pm"), 
                           1, 0)) |> 
  filter(is_anwar == 1) |> 
  filter(!str_detect(body, "Users often report submissions|I am a bot")) |> 
  mutate(body = str_replace_all(body, "non |Non ", "non-")) |> 
  mutate(score_trend = case_when(
    score > 1 ~ "positive", 
    score == 1 ~ "neutral", 
    score > 10 ~ "very positive", 
    score <= 0 ~ "negative"
  )) |> 
  select(comment_id, thread_id, body, score, score_trend, flair) |> 
  mutate(body = remove_url(body)) |> 
  unnest_tokens(bigram, body, token = "ngrams", n = 2) |> 
  add_count(bigram) |> 
  filter(!is.na(bigram)) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  mutate(word1 = ifelse(word1 %in% malay_slang$slang, malay_slang$translation, word1)) |>
  mutate(word2 = ifelse(word2 %in% malay_slang$slang, malay_slang$translation, word2)) |>
  filter(!word1 %in% stop_words$word & str_detect(word1, "[a-z]")) |> 
  filter(!word2 %in% stop_words$word & str_detect(word2, "[a-z]")) |> 
  count(word1, word2, thread_id, sort = TRUE) |> 
  unite(bigram, word1, word2, sep = " ") |> 
  bind_tf_idf(bigram, thread_id, n)


```


```{r}
anwar_bigrams_dtm <- anwar_bigrams |> 
  cast_dtm(thread_id, bigram, n)

anwar_bigrams_lda <- 
  LDA(anwar_bigrams_dtm, k = 4, control = list(seed = 133))

anwar_bigrams_topics <- tidy(anwar_bigrams_lda, matrix = "beta")
```



```{r}
anwar_bigrams_topics |> 
  filter(term %out% c("gif giphy", "subreddit message", "performed automatically", "message compose", 
                      "img.emote", "img emote", "collecting responses", "march till", 
                      "prime minister", "anwar ibrahim", "match till", 
                      "report submissions", "removed due", "sensationalised articles", 
                      "check sources", "unity government", "madani government", 
                      "datuk seri", "skeptical check", "reduce spam", "parent commenter", 
                      "misleading contact", "article's claims", "social media")) |> 
  filter(!str_detect(term, "11.59pm|23rd|emote|26th|blanket")) |> 
  group_by(topic) |> 
  slice_max(beta, n = 20) |> 
  ungroup() |> 
  arrange(topic, -beta) |> 
  ggplot(aes(x = beta, y = reorder_within(term, beta, topic), 
             fill = topic)) + 
  geom_col() + 
  facet_wrap(~ topic, scales = "free_y") + 
  scale_y_reordered()
```
```{r}
comments |> 
  filter(str_detect(body, "ini semua")) |> 
  sample_n(5) |> 
  pull(body)
```


```{r}
google_rating_bigrams |> 
  mutate(rating = as.factor(rating)) |> 
  arrange(desc(n)) |> 
  group_by(rating) |> 
  slice(1:15) |> 
  ungroup() |> 
  ggplot(aes(x = n, y = reorder_within(bigram, n, rating))) + 
  geom_col(aes(fill = rating))

malaystopwords |> glimpse()
malayrootwords |> glimpse()
```


```{r}

google_rating_words <- google_kl_cuisine |> 
  select(rowid, rating, review) |> 
  unnest_tokens(word, review) |> 
  # mutate(word = SnowballC::wordStem(word, language = "porter")) |> 
  anti_join(stop_words, by = "word") |>
  add_count(word, rating) 


  
  
  mutate(count = 1) |> 
  mutate(
    bins = case_when(
    score < 0 & score >= -10 ~ "0_-10",
    score >= 0 & score <= 10 ~ "0_10", 
    score > 10 & score <= 30 ~ "11_30", 
    score > 30 ~ ">30", 
    TRUE ~ "<-10"),
    bins = fct_relevel(bins, 
                       c("<-10", 
                         "0_-10",
                         "0_10", 
                         "11_30",
                         ">30"))) |>
  ggplot(aes(x = count, y = bins)) + 
  geom_col()
  geom_histogram() + 
  scale_x_continuous(trans = "pseudo_log", 
                     breaks = c(-200, -100, -30, 0, 30, 100, 300, 500, 1000))
  count(score)
  count(is_anwar)
```

