---
title: "Malaysian Elections"
subtitle: "Part two: Ethnicity and Election Predictions"
author: "Sean Ng"
organization: "AIMdata"
date-modified: "11 January 2026"
execute: 
  echo: false
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 9)


library(tidyverse)
library(here)
library(janitor)
library(scales)
library(tidytext)
library(widyr)
library(ggraph)
library(patchwork)
library(kableExtra)
library(viridis)
library(stringr)
library(sf)
library(ggmagnify)
library(ggthemes)
library(glmnet)
library(caret)
library(plotly)

`%out%` <- Negate(`%in%`)
options(scipen = 100)
theme_set(theme_light())
range_wna <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

last_comma <- function(x, sep = ","){
  x <- sapply(strsplit(x, sep),"[")
  trimws(x[[length(x)]])
}
```




```{r data}

census_dun <- read_csv("./data/census_dun.csv") |> 
  mutate(population_bumi = ethnicity_proportion_bumi * population_total / 100, 
         population_chinese = ethnicity_proportion_chinese * population_total/ 100, 
         population_indian = ethnicity_proportion_indian * population_total / 100, 
         population_other = ethnicity_proportion_other * population_total / 100)


gini <- read_csv("./data/hh_inequality_parlimen.csv") |> 
  filter(date == "2024-01-01")


sf::sf_use_s2(FALSE)

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))

federal_territories <- tribble(
  ~parlimen, ~population_total, ~pc_bumi, ~pc_chinese, ~pc_indian, ~pc_other, ~sex_male, ~sex_female, ~population_density, ~income_avg, ~expenditure_avg, ~poverty_incidence, ~area_km2,
  "P.114 Kepong", 106199, .202, .729, .063, .006, .588, .412, 8850, 13087, 6500, .003, 12,
  "P.115 Batu", 219132, .566, .301, .129, .004, .601, .399, 10957, 11019, 5817, 0, 20,
  "P.116 Wangsa Maju", 215600, .617, .306, .07, .008, .537, .463, 13475, 11153, 5768, 0, 16,
  "P.117 Segambut", 253715, .324, .558, .107, .012, .512, .479, 4975, 20521, 9845, 0, 51,
  "P.118 Setiawangsa", 147095, .697, .214, .085, .004, .519, .481, 9193, 12281, 6534, .003, 16, 
  "P.119 Titiwangsa", 122096, .677, .233, .076, .014, .519, .481, 8140, 11921, 6823, .005, 15,
  "P.120 Bukit Bintang", 120529, .271, .565, .155, .009, .565, .435, 5739, 13677, 7401, 0, 21, 
  "P.121 Lembah Pantai", 148094, .591, .262, .138, .009, .509, .491, 7405, 13081, 6186, 0, 20,
  "P.122 Seputeh", 322511, .311, .573, .112, .004, .510, .490, 10404, 13489, 7764, .005, 31,
  "P.123 Cheras", 135823, .322, .596, .076, .007, .527, .473, 8489, 12031, 5979, .002, 16,
  "P.124 Bandar Tun Razak", 191318, .679, .234, .083, .004, .509, .491, 7653, 11266, 6002, 0, 25,
  "P.166 Labuan", 95120, .862, .117, .011, .01, .514, .486, 1034, 8319, 4097, .031, 96,
  "P.125 Putrajaya", 109202, .979, .012, .006, .003, .509, .491, 2229, 12840, 7980, .004, 49
  
)

census_fed <- census_dun |> 
  mutate(population_poor = poverty_incidence * population_total / 100) |> 
  group_by(parlimen) |> 
  summarise_at(
    c("population_total", 
      "population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ sum(.x, na.rm = TRUE)
  ) |> 
  mutate_at(
    c("population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ .x / population_total
  ) |>  
  rename(
    pc_bumi = population_bumi, 
    pc_chinese = population_chinese, 
    pc_indian = population_indian, 
    pc_other = population_other, 
    poverty_incidence = population_poor) |> 
  left_join(
    census_dun |>
      mutate(
        income = income_avg * household_total,
        expenditure = expenditure_avg * household_total) |>
      group_by(parlimen) |>
      summarise(
        household_total = sum(household_total),
        income = sum(income),
        expenditure = sum(expenditure)) |>
      mutate(
        income_avg = income / household_total,
        expenditure_avg = expenditure / household_total) |>
      select(parlimen,
             income_avg, 
             expenditure_avg), 
    by = "parlimen") |> 
  left_join(
    parlimen_geo |> 
      st_drop_geometry() |> 
      select(parlimen, area_km2), 
    by = "parlimen"
  ) |> 
  mutate(population_density = population_total / area_km2) |> 
  rbind(federal_territories) |> 
  left_join(
    census_dun |> 
      distinct(parlimen, state), 
    by = "parlimen"
  ) |>
  left_join(
    gini |> 
      select(parlimen, gini), 
    by = "parlimen"
  ) |> 
  mutate(state = case_when(
    is.na(state) & str_detect(parlimen, "Labuan") ~ "Labuan", 
    is.na(state) & str_detect(parlimen, "Putrajaya") ~ "Putrajaya", 
    is.na(state) ~ "Kuala Lumpur", 
    TRUE ~ state
  )) |> 
  mutate(east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                                1, 0))

ballots <- read_csv("./data/consol_ballots.csv") |> 
  mutate(
    age = ifelse(age == -1, NA_real_, age)
  ) |> 
  mutate(federal = ifelse(
    str_detect(seat, "P.0|P.1|P.2"), 
    "Federal", 
    "State"))

stats <- read_csv("./data/consol_stats.csv")

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))

multiracial_races <- ballots |> 
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(count = 1) |> 
  # I think this filter is fine
  filter(rank <= 3) |> 
  group_by(state, seat, date, federal, election) |> 
  summarise(malay_candidates = sum(count[ethnicity == "Malay"]), 
            chinese_candidates = sum(count[ethnicity == "Chinese"]), 
            indian_candidates = sum(count[ethnicity == "Indian"]), 
            sabah_candidates = sum(count[ethnicity == "Bumi Sabah"]), 
            sarawak_candidates = sum(count[ethnicity == "Bumi Sarawak"]), 
            other_candidates = sum(count[ethnicity == "Other"]), 
            .groups = "drop") |> 
  mutate(total_candidates = malay_candidates + chinese_candidates + indian_candidates + 
           sabah_candidates + sarawak_candidates + other_candidates) |> 
  pivot_longer(cols = malay_candidates:other_candidates, 
               names_to = "candidate_ethnicity", 
               values_to = "value") |> 
  mutate(candidate_ethnicity = str_remove_all(candidate_ethnicity, "_candidates")) |> 
  filter(value != 0) |>
  mutate(multiracial = ifelse(value != total_candidates, 1, 0)) |> 
  mutate(combination = ifelse(
    multiracial == 0, paste0(candidate_ethnicity, "_only"), "multiethnic"
  )) |> 
  distinct(state, seat, date, federal, election, multiracial, combination)
```

**Acronyms**
```{r table-acronyms, echo = FALSE, message=FALSE, warning=FALSE}
tribble(
  ~acronym, ~name, 
  "BN", "Barisan Nasional",  
  "DAP", "Democratic Action Party",
  "GPS", "Gabnungan Parti Sarawak",
  "GRS", "Gabungan Rakyat Sabah", 
  "GS", "Gagasan Sejahtera", 
  "MCA", "Malaysian Chinese Association", 
  "MIC", "Malaysian Indian Congress", 
  "PAS", "Parti Islam Se-Malaysia", 
  "PH", "Pakatan Harapan", 
  "PKR", "Parti Keadilan Rakyat", 
  "PN", "Perikatan Nasional", 
  "SUPP", "Sarawak United People's Party", 
  "UMNO", "United Malays National Organisation", 
  "USA", "United Sabah Alliance"
) |>  
  kable(col.names = NULL, format = "html") |> 
  kable_classic(bootstrap_options = c("condensed"), 
                full_width = FALSE, 
                position = "float_left")
```

<br>

# Racialisation of elections

According to an [Architects of Diversity survey](https://www.aodmalaysia.org/sods), 64% of Malaysians reporting experiencing some for of discrimination in the past 12 months, with 38% saying the discrimination was related to socioeconomic status, 33% saying it was related to age and 32% saying it was related to ethnicity. 

Hindus were the most likely to have experienced religious-based discrimination (40%), compared to their Christian (26%), Buddhists (22%) and Muslims (20%). 59% of Muslim respondents believe their own religious group faced a lot discrimination, with the figure being 63% amongst Buddhists, 76% amongst Christians and 81% amongst Hindus. 

According to [Pusat KOMAS](https://komas.org/download/Malaysia-Racism-Report_A-Decade-In-Review-2015-2024.pdf), election years tend to see "sharp increases in racial discrimination narratives [...] Political parties have consistently used race and religion as mobilization tools". The add, "politics accounted for just over 43% of all recorded racial discrimination incidents between 2015 and 2024" and that "racial and religious politics remain the primary diver of racial discrimination in Malaysia". 

The next general election is likely to witness another surge in racial politics, particularly with PAS and Bersatu continuing to consolidate their Malay-Muslim voter bases. 

<br><br><br>


```{r }

parlimen_geo |> 
  left_join(
    multiracial_races |>
      filter(federal == "Federal") |>
      select(seat, multiracial, combination) |>
      group_by(seat) |>
      slice_max(order_by = multiracial) |>
      ungroup() |>
      distinct(seat, multiracial, combination), 
    by = c("parlimen" = "seat") 
  ) |> 
  mutate(multiracial = ifelse(multiracial == 1, "Multiethnic", "Single ethnicity")) |> 
  ggplot() + 
  geom_sf(aes(fill = multiracial), 
          linewidth = .01, 
          colour = "grey70") + 
  scale_fill_viridis_d(option = "magma", 
                       begin = .2, 
                       end = .9) + 
  # Only one inset can work at one time
  # Solution proposed here https://github.com/hughjonesd/ggmagnify/issues/30 
  # Does not work
  geom_magnify(
    from = c(101.223422,  101.953542, 2.805661, 3.383667), 
    to = c(105, 108, 3.5, 5.9), 
    shadow = FALSE,
    proj = "single"
  ) + 
    #theme_map() + 
  labs(
    fill = "",
    title = "Federal constituencies which have ever had multiethnic electoral races"
  ) + 
  theme_void() +

parlimen_geo |> 
  left_join(
    census_fed |>
      select(parlimen, pc_bumi), 
    by = "parlimen"
  ) |> 
  ggplot() + 
  geom_sf(aes(fill = pc_bumi), 
          linewidth = .01, 
          colour = "grey70") + 
  scale_fill_viridis(option = "magma", 
                       begin = .2, 
                       end = .9, 
                     labels = percent, 
                     breaks = seq(.1, .9, .1)) + 
  # Only one inset can work at one time
  # Solution proposed here https://github.com/hughjonesd/ggmagnify/issues/30 
  # Does not work
  geom_magnify(
    from = c(101.223422,  101.953542, 2.805661, 3.383667), 
    to = c(105, 108, 3.5, 5.9), 
    shadow = FALSE,
    proj = "single"
  ) + 
    #theme_map() + 
  labs(
    fill = "",
    title = "Federal constituencies by the proportion of their population who are Bumiputera"
  ) + 
  theme_void() +
  
  plot_layout(ncol = 1)
  
```

<br>

43% (97 out of 222 constituencies) have ever had multiethnic electoral races. These constituencies align largely with the areas with the highest concentrations of non-Bumiputeras. 



<br>

```{r}
ballots |>
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1) |>
  # This is just to isolate the number of electoral races
  filter(result %in% c("won", "won_uncontested")) |> 
  group_by(
    decade, federal, combination
  ) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count), 
         combination = case_when(
           combination == "sabah_only" ~ "bumi_sabah_only", 
           combination == "sarawak_only" ~ "bumi_sarawak_only", 
           TRUE ~ combination), 
         combination = str_to_title(str_replace_all(combination, "_", " "))
         ) |> 
  ungroup() |>
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, y = pc, fill = fct_relevel(combination, 
                                   c("Bumi Sabah Only", 
                                     "Bumi Sarawak Only", 
                                     "Chinese Only", 
                                     "Indian Only", 
                                     "Malay Only", 
                                     "Other Only", 
                                     "Multiethnic")))) + 
  geom_col() + 
  facet_wrap(~ federal) + 
  scale_y_continuous(labels = percent) +
  scale_fill_manual(
    values = c(
      "Bumi Sabah Only" = "#440154FF", 
      "Bumi Sarawak Only" = "#414487FF",
      "Chinese Only" = "#2A788EFF",
      "Indian Only" = "#22A884FF",
      "Malay Only" = "#7AD151FF",
      "Other Only" = "#FDE725FF",
      "Multiethnic" = "#FCFDBFFF"
    )
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(fill = "Combination",
       title = "Multiethnic electoral races were least common during the 90s and 00s", 
       x = "", 
       y = "") + 
  theme(strip.background = element_rect(fill = "grey20"), 
        axis.text.x = element_text(size = 7))
```
<br>

Most constituencies in Malaysia are multi-ethnic in nature: `r census_fed |> filter(pc_bumi > .75) |> nrow()` out of the `r census_fed |> nrow()` federal constituencies have at least 25% of their populations being racial minorities. The latest redelineation was done in 2006 and the plots in this section will focus on the years since then. 

Commonsensically, given the racialised nature of Malaysian politics, Bumiputera candidates are far more likely to win in areas that have high proportions of Bumiputera candidates. Non-bumiputera candidates almost never win seats in constituencies with more than 75% Bumiputera residents. 

<br> 

```{r}
ballots |>
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1, 
         ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |>
  filter(result %in% c("won", "won_uncontested") & combination == "multiethnic") |> 
  group_by(
    decade, federal, combination, ethnicity
  ) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count)) |> 
  ungroup() |>
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, y = pc, fill = fct_relevel(ethnicity, 
                                   c("Bumi Sabah", 
                                     "Bumi Sarawak", 
                                     "Chinese", 
                                     "Indian", 
                                     "Malay", 
                                     "Other")))) + 
  geom_col() + 
  facet_wrap(~ federal) + 
  scale_y_continuous(labels = percent) +
  scale_fill_manual(
    values = c(
      "Bumi Sabah" = "#440154FF", 
      "Bumi Sarawak" = "#414487FF",
      "Chinese" = "#2A788EFF",
      "Indian" = "#22A884FF",
      "Malay" = "#7AD151FF",
      "Other" = "#FDE725FF")
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(fill = "Combination",
       title = "Multiethnic electoral races were most commonly won by Chinese candidates",
       subtitle = "Though this has been on the downtrend since the 00s",
       x = "", 
       y = "") + 
  theme(strip.background = element_rect(fill = "grey20"), 
        axis.text.x = element_text(size = 7))
```


<br>

From the plot below, we also see that Bumiputeras tend only to win when a constituency has a population that is more than 50% Bumi. 

<br>


```{r}
ballots |>
  filter(date > "2006-01-01") |> 
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  left_join(
    census_fed |> 
      select(parlimen, pc_bumi), 
    by = c("seat" = "parlimen")
  ) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  mutate(
    bumiputera = ifelse(ethnicity %in% c("Malay", "Bumi Sabah", "Bumi Sarawak"),
                        "Bumiputera winner",
                        "non-Bumi winner"),
    east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                           "East Malaysia", "Peninsula")) |> 
  ggplot(aes(x = pc_bumi, fill = east_malaysia)) + 
  geom_histogram() + 
  scale_fill_viridis_d(begin = .1, end = .9) +
  scale_x_continuous(labels = percent) +
  facet_wrap(~ bumiputera) + 
  labs(
    title = "Electoral races by % of the constituency who are Bumiputera", 
    subtitle = "Shows races since the 2006 redelineation", 
    fill = "", 
    y = "Number of electoral races", 
    x = "% of constituency who are Bumiputera"
  ) + 
  theme(strip.background = element_rect(fill = "grey20"))

```

<br>

The coalition which benefited primarily from this racialised dimension to elections has been Barisan Nasional. 

Almost all of BN's electoral wins come from constituencies which are more than 75% Bumiputera. This is the same for Perikatan Nasional. Conversely, constituencies which are more than 75% Bumiputera almost never vote for parties under the Pakatan Harapan (or its predecessor Pakatan Rakyat) coalition. 

<br>

```{r}
ballots |>
  filter(date > "2006-01-01") |> 
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  left_join(
    census_fed |> 
      select(parlimen, pc_bumi), 
    by = c("seat" = "parlimen")
  ) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  mutate(
    bumiputera = ifelse(ethnicity %in% c("Malay", "Bumi Sabah", "Bumi Sarawak"),
                        "Bumiputera winner",
                        "non-Bumi winner"),
    east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                           "East Malaysia", "Peninsula")) |> 
  ggplot(aes(x = pc_bumi, fill = coalition)) + 
  geom_histogram() + 
  scale_fill_manual(values = c(
    "ALONE" = "grey50", 
    "BN" = "#1D1147FF", 
    "GPS" = "#51127CFF", 
    "GRS" = "#822681FF", 
    "GS" = "#B63679FF", 
    "PH" = "#E65164FF", 
    "PN" = "#FB8861FF", 
    "PR" = "#FEC287FF", 
    "USA" = "#FCFDBFFF"
  )) +
  scale_x_continuous(labels = percent) +
  facet_wrap(~ bumiputera) + 
  labs(
    title = "Electoral races by % of the constituency who are Bumiputera", 
    subtitle = "Shows races since the 2006 redelineation", 
    fill = "Coalition", 
    y = "Number of electoral races", 
    x = "% of constituency who are Bumiputera"
  ) + 
  theme(strip.background = element_rect(fill = "grey20"))
```

<br> 


A significant minority of non-Bumiputera MPs were from BN. Additionally, most of those wins came from constituencies with higher Bumiputera populations, indicating that party loyalty (which itself has racial connotations) were stronger than the ethnicity of individual candidates. 

<br><br><br>

# Ethnic makeup of parties and coalitions

Amongst the various coalitions, East Malysian Coalitions and Gagasan Sejahtera were the most racially homogenous. Proportionally, Pakatan Harapan and Pakatan Rakyat both fielded more Malay candidates than Barisan Nasional. 

However, as noted by [Mustafa, Kartini and Nazri](https://www.frontiersin.org/journals/political-science/articles/10.3389/fpos.2026.1721966/full), whilst BN might be multiethnic in nature, the coalition was not consociational, but was "sustained more by UMNO’s control and patronage than by genuine power-sharing". 


<br>

```{r}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |> 
  filter(rank <= 3) |>
  group_by(coalition, ethnicity) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(coalition) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates) |> 
  ggplot(aes(x = pc, y = fct_rev(coalition))) + 
  geom_col(aes(fill = ethnicity)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of candidates", 
       y = "Coalition", 
       title = "Candidates fielded by ethnicity", 
       subtitle = "Only showing politicians who ran since the redelineation in 2006", 
       fill = "Ethnicity")
  
  
```

<br>

The plot below shows the breakdown of candidates by ethnicity and party. No party reflects the racial make-up of the country. Amanah, Bersatu, PAS and UMNO are Malay parties, DAP, Gerakan and the MCA are Chinese parties. The MIC is an Indian party, PBB is a Sabahan party and WARISAN is a Sarawakian party. 

When we look at party level, we note that BEBAS and PKR are the only true multiracial parties, with all other parties being dominated by a main ethnic group, which only underscores the heavy racial component of Malaysian politics. 

<br> 


```{r ethnicity-party-candidates}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |>
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(party = fct_lump(party, n = 12)) |> 
  group_by(party, ethnicity) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(party) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates) |> 
  ggplot(aes(x = pc, y = fct_rev(fct_relevel(party, 
                                             c("Other", "UMNO", "PAS", "PKR", 
                                               "BEBAS", "DAP", "MCA", "BERSATU", 
                                               "AMANAH", "WARISAN", "PBB","GERAKAN", "STAR"
                                             ))))) + 
  geom_col(aes(fill = ethnicity)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of candidates", 
       title = "The only true multi-ethnic parties are BEBAS and PKR", 
       subtitle = "Only candidates since 2006; parties ordered by number of candidates", 
       fill = "Candidate\nethnicity")
  
  
```

<br>


These patterns hold when we examine the candidates who actually won. The plot below shows the breakdown of MPs by party since 2006. 


<br>

```{r}
ballots |> 
  filter(date > "2006-01-01") |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  mutate(count = 1) |>
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(party = fct_lump(party, n = 12)) |> 
  group_by(party, ethnicity) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(party) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates) |> 
  ggplot(aes(x = pc, y = fct_rev(fct_relevel(party, 
                                     c("UMNO", "DAP", "PAS", "PKR", "PBB", 
                                       "Other", "BERSATU", "WARISAN", "MCA", 
                                       "AMANAH", "PRS", "PBS", "SUPP"))))) + 
  geom_col(aes(fill = ethnicity)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of MPs", 
       title = "The only true multi-ethnic parties are BEBAS and PKR", 
       subtitle = "Since 2006; parties sorted by number of MPs", 
       fill = "Candidate\nethnicity")
  

```

<br><br><br>

# The collapse of UMNO and Islamic politics

According to [Deivasagayam](https://msocialsciences.com/index.php/mjssh/article/view/3248/2134), the two main cleavages in Malaysian society -- race and religion -- have converged, contributing to UMNO's downfall. He continues, "UMNO consistently positioned itself as the exclusive advocate and guardian of Malay rights and privileges". However, by engaging in an "Islamisation Race" to stave off PAS, UMNO inadvertently legitimised PAS in the eyes of Muslim voters, and when it advocated for Malay Islamic Supremacy, it set that stage for the convergence for race and religion which would later ultimately benefit PAS and BERSATU, which demonstrated the possibility of a Malay-Muslim governance structure which excluded minorities. 


We see, from the plot below, the extent of UMNO's electoral collapse. From its peak of `r filter(ballots, election == "GE-11" & result %in% c("won", "won_uncontested") & party == "UMNO") |> nrow()` seats under Badawi, it now holds less seats than either PAS, DAP, PKR or Bersatu. 

<br>


```{r parties-mp-share}
ballots |> 
  mutate(count = 1) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  filter(str_detect(election, "GE")) |> 
  mutate(year = year(date)) |> 
  group_by(election) |> 
  mutate(year = min(year)) |> 
  ungroup() |> 
  mutate(party = fct_lump(party, n = 12, w = votes)) |> 
  group_by(year, party) |> 
  summarise(mps = sum(count), 
            .groups = "drop") |> 
  ggplot(aes(x = year, y = mps, group = party)) + 
  geom_line(aes(colour = party), 
            linewidth = 1.05, 
            alpha = .8) + 
  # scale_colour_manual(
  #   values = c(
  #     "UMNO" = "#30123BFF",
  #     "MCA" = "#4454C4FF",
  #     "MIC" = "#4490FEFF",
  #     "BERSATU" = "#1FC8DEFF",
  #     "PAS" = "#29EFA2FF",
  #     "GERAKAN" = "#7DFF56FF",
  #     "PBB" = "#C1F334FF",
  #     "PBS" = "#F1CA3AFF",
  #     "USNO" = "#FE922AFF",
  #     "SUPP" = "#EA4F0DFF",
  #     "PKR" = "#BE2102FF",
  #     "DAP" = "#7A0403FF",
  #     "Other" = "grey50"
  #     
  #   )
  # ) +
  scale_colour_viridis_d(option = "turbo") +
  scale_x_continuous(breaks = seq(1955, 2026, 5)) +
  scale_y_continuous(breaks = seq(0, 120, 10)) +
  guides(colour = guide_legend(override.aes = list(linewidth = 2, 
                                                   alpha = 1))) + 
  labs(x = "", 
       y = "Number of MPs", 
       title = "The collapse of UMNO has left PAS and DAP as the largest parties", 
       colour = "") + 
  theme(axis.text.x = element_text(size = 7))
```
<br>

However, when we examine their share of votes, we see that their long dominance of parliament was always more fragile than was perceived. At no time after 1960 did UMNO ever win more than 40% of the votes. Additionally, the Badawi peak of `r filter(ballots, election == "GE-11" & result %in% c("won", "won_uncontested") & party == "UMNO") |> nrow()` seats was accomplished with only 35% of the national vote, indicating their reliance on a first-past-the-post system. Though now, both UMNO and PKR are penalised by the first-past-the-post system, whereas PAS, DAP and Bersatu are each rewarded by it. 


Finally, though their share of votes is lower than it has ever been, UMNO still, marginally, received more votes than any other party. 

<br> 

```{r parties-vote-share}
ballots |>  
  filter(str_detect(election, "GE")) |>
  mutate(year = year(date)) |> 
  group_by(election) |> 
  mutate(year = min(year)) |>
  ungroup() |> 
  mutate(party = fct_lump(party, n = 12, w = votes)) |>
  group_by(year, party) |> 
  summarise(votes = sum(votes), .groups = "drop") |>
  group_by(year) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = year, y = votes_pc, group = party)) + 
  geom_line(aes(colour = party), 
            linewidth = 1.05, 
            alpha = .8) + 
  scale_colour_viridis_d(option = "turbo") +
  scale_x_continuous(breaks = seq(1955, 2026, 5)) +
  guides(colour = guide_legend(override.aes = list(linewidth = 2, 
                                                   alpha = 1))) + 
  scale_y_continuous(labels = percent, breaks = seq(0, .6, .1)) +
  labs(x = "", 
       y = "Share of popular vote", 
       title = "UMNO still has, very marginally, the largest share of votes",
       subtitle = "But it has never won more than 40% of votes since the formation of the Federation", 
       colour = "") + 
  theme(axis.text.x = element_text(size = 7))
  
  
```

<br> 

Recalling Deivasagayam's arguments, [ISEAS](https://www.iseas.edu.sg/wp-content/uploads/2024/04/TRS12_24.pdf) concludes that "the sense of Islamic identity and belonging often supersedes the allure of economic development for Malay youths when choosing how to cast their vote". 

This speaks to not only to the Islamisation race that UMNO, PAS and Bersatu have engaged in: 65% of respondents in Muslim Youth Survey saying that Muslims should only vote for Muslim leaders, but also to an unfair economic system where whatever economic development is meagre and piecemeal, and insufficient to attract voters.


Recalling our earlier criticism of Anwar, is his bid with GRS meant to engineer a more permanent parliamentary majority without needing to win the popular vote? 

<br><br><br>


 

Pakatan Harapan seems unwilling to engage in the Islamisation Race or put forward a competing vision of a more moderate Islam, possibly out of fears that doing so would alienate its core non-Muslim base and anger conservative Muslims.

There is a more revealing dichotomy between Malay and non-Malay voters’ perception towards the government which points to an enduring ethnocentric political choice. The survey found that 85% of Malay voters felt that the PN government protects the interest of the Malays, while 77% of non-Malay voters believed the government does not treat all races equally. This is key to understanding political choices among different ethnic groups in Malaysia, not only in the 15th General Election, but also in previous elections.

Is this why the malapportionment deal with GRS is so important? Because Pakatan Harapan cannot envision itself winning over a majority of Malay voters? 



Be that as it may, a closer look at BN and PN leaders’ approval rating among Malay voters reveals a slightly different scenario. PN Chairman and former Prime Minister, Muhyiddin Yassin, received a higher approval rating (67%) compared to BN Chairman, Zahid Hamidi (19%).


 
# Can you win with only Malay voters? 

The PN/BN government did prove

<br><br><br>

# Fragile coalitions

Whilst public support for BN has faded considerably, between BN and PN, there was majority support for a Malay-supremacist (*Ketuanan Melayu*) government which collected more than 50% of votes in GE-15, an increase in the vote share garnered by BN and GS in GE-14. 

The current unity (*Madani*) government is heavily reliant on BN for both a parliamentary majority as well as a majority of the vote share. 

<br>

```{r general-elections-vote-share}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA"), "Other", coalition
  )) |> 
  filter(election %in% c("GE-15", "GE-14", "GE-13", "GE-12")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            candidates = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = votes_pc, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(labels = percent) + 
  labs(x = "% of votes", 
       y = "", 
       fill = "Coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br>

The two largest coalitions in parliament are now PH and PN. But since neither were able to win a commanding mandate in GE-15, BN was left to play kingmaker. 

<br>

```{r general-elections-seats}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA"), "Other", coalition
  )) |> 
  filter(election %in% c("GE-15", "GE-14", "GE-13", "GE-12") & 
           result %in% c("won", "won_uncontested")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            seats = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = seats, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(breaks = seq(0, 150, 20)) +
  labs(x = "Parliament seats", 
       y = "", 
       fill = "Coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br>

The plots below show vote share amongst the major coalitions by state elections. We see that that PN has managed to garner a lead over PH in state polls. 

Perhaps this is what Anwar's agreement with GRS to increase the number of seats in Sabah is about. Though, we still emphasise that any further malapportionment would be, ultimately, an anti-democratic exercise. 

<br>

```{r state-elections-coalitions}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA", "WARISAN-PLUS", "GASAK"), "Other", coalition
  )) |> 
  filter(election %in% c("SE-15", "SE-14", "SE-13", "SE-12")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            candidates = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = votes_pc, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(labels = percent) + 
  labs(x = "% of votes", 
       y = "", 
       fill = "Coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br><br><br>

# Predictions for the GE-16


```{r training-and-testing}

training <- census_fed |> 
  mutate(pc_bumi_peninsula = ifelse(
    east_malaysia == 0, pc_bumi, 0), 
    pc_bumi_east_malaysia = ifelse(
      east_malaysia == 1, pc_bumi, 0
    )) |> 
  select(
    parlimen, population_density, pc_bumi_peninsula, pc_bumi_east_malaysia, 
    pc_chinese, pc_indian, gini, income_avg
  ) |> 
  left_join(
    ballots |>
      filter(election == "SE-15") |>
      rbind(ballots |>
              filter(date == "2021-12-18" & state == "Sarawak")) |>
      left_join(census_dun |>
                  distinct(dun, parlimen), by = c("seat" = "dun")) |>
      group_by(parlimen) |>
      summarise(
        ph_votes = sum(votes[coalition == "PH"], na.rm = TRUE),
        total_votes = sum(votes, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(pc_ph_votes = ph_votes / total_votes), 
    by = "parlimen"
  ) |> 
  filter(!is.na(pc_ph_votes)) |>
  mutate(population_density = log(population_density)) |> 
  mutate_at(
    c("population_density", "pc_bumi_peninsula", "pc_bumi_east_malaysia", 
      "pc_chinese", "pc_indian", "gini", "income_avg"), 
    function(x){
      (x - mean(x) / sd(x))
    }
  ) |> 
  na.omit()

testing <- census_fed |> 
  mutate(pc_bumi_peninsula = ifelse(
    east_malaysia == 0, pc_bumi, 0), 
    pc_bumi_east_malaysia = ifelse(
      east_malaysia == 1, pc_bumi, 0
    )) |> 
  select(
    parlimen, population_density, pc_bumi_peninsula, pc_bumi_east_malaysia, 
    pc_chinese, pc_indian, gini, income_avg
  ) |> 
  left_join(
    ballots |> 
      filter(election == "GE-15") |> 
      group_by(seat) |> 
      summarise(ph_votes = sum(votes[coalition == "PH"], na.rm = TRUE), 
                total_votes = sum(votes, na.rm = TRUE), 
                .groups = "drop") |>
      mutate(pc_ph_votes = ph_votes / total_votes), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_density = log(population_density)) |> 
  mutate_at(
    c("population_density", "pc_bumi_peninsula", "pc_bumi_east_malaysia", 
      "pc_chinese", "pc_indian", "gini", "income_avg"), 
    function(x){
      (x - mean(x) / sd(x))
    }
  ) |> 
  na.omit()
```


```{r stepwise-model}
stepwise_model <- step(lm(
  pc_ph_votes ~ population_density + pc_bumi_peninsula + pc_bumi_east_malaysia + 
  pc_chinese + pc_indian + gini + income_avg, 
  data = training, 
  direction = "backward"
))

# The stepwise model indicates that the most relevant variables to predicting 
# PH performance are log(population density), % of population who are peninsula bumis, 
# % of the population who are east malaysia bumis and % of the population who are indian
```


```{r glmnet-model}
glmnet_model <- train(
  pc_ph_votes ~ population_density + pc_bumi_peninsula + pc_bumi_east_malaysia + 
  pc_chinese + pc_indian + gini + income_avg, 
  data = training, 
  method = "glmnet", 
  metric = "RMSE", 
  trControl = trainControl(method = "LOOCV"), 
  preProcess = c("center", "scale"), 
  tuneLength = 10
)

```

```{r preds}
preds <- testing %>%
  mutate(
    aic_pred = predict(object = stepwise_model, newdata = .),
    glmnet_pred = predict(object = glmnet_model, newdata = testing)
  ) %>%
  mutate(pred = (aic_pred + glmnet_pred) / 2) %>%
  pull(pred)

testing$proj_ph_vote <- preds
```

In this section, we will try and predict Pakatan Harapan's performance (in this case, the percentage of votes we anticipate them to receive) in the upcoming general election. We have combined predictions from a stepwise model and a glmnet model below. 

<br>

```{r plotting-results}
pred_plot <- testing |> 
  left_join(
    ballots |> 
      filter(election == "GE-15" & result %in% c("won", "won_uncontested")) |> 
      select(state, seat, coalition), 
    by = c("parlimen" = "seat")
  ) |> 
  ggplot(aes(x = pc_ph_votes, y = proj_ph_vote)) + 
  geom_point(aes(colour = coalition,
                 text = paste0(parlimen, ",", "\n", 
                               state, ",", "\n", 
                               "Projected PH%: ", round(proj_ph_vote * 100, 2), "%", "\n", 
                               "GE-15 PH%: ", round(pc_ph_votes * 100, 2), "%", "\n", 
                               "GE-15 winner: ", coalition))) +
  geom_smooth(method = "lm") + 
  scale_colour_viridis_d(option = "turbo") + 
  scale_x_continuous(labels = percent, breaks = seq(0, 1, .1)) + 
  scale_y_continuous(labels = percent, breaks = seq(0, 1, .1)) +
  labs(colour = "Winning\ncoalition\nin GE15", 
       x = "% PH votes in GE-15", 
       y = "Projected % of PH votes in GE-16")

ggplotly(pred_plot, tooltip = c("text")) |> 
  layout(title = list(text = 
                        paste0(
                          "Interactive plot of GE-15 results and predictions for GE-16", 
                          "<br>",
                          "<sup>", "mouse over for details; drag and click to select and zoom; double-click legend select/deselect", "</sup>"
                        )))

```


<br>

We see the extent of the instability of the Madani's government's coalition from the plotted predictions above. Marginal seats where PH may seek to expand its reach will largely come at the expense of BN and vice versa.

There are a fair number of marginal seats where PH received less than 40% of the votes in GE-15, but were still won by PH (many of these were three-cornered fights). These are all vulnerable for pickup by BN or PN. 

PH may extend their reach by running better races and fielding better candidates in Bagan Datuk in Perak, Titiwangsa in KL, Tampin and Jempol in Negeri Sembilan, Kuala Langat in Selangor and Padang Serai in Kedah. 

We finally also note that the model's predictions are not particularly reliable when it comes to seats in East Malaysia and a tendency to undercount PH's margin. 


## Three-cornered fights

However, given the loose alliances between the various coalitions, three-cornered fights have become much more prevalent. 

How do we account for three-cornered fights? And what is their impact on PH? 
How to factor in three-cornered fights into the model? 




```{r}
ballots |> 
  # Beginning of the collapse of BN and the opening up of 
  # the political space
  filter(date > "2013-01-01" & votes_perc >= 5) |> 
  mutate(count = 1) |> 
  group_by(seat, date, federal) |> 
  summarise(candidates = sum(count), .groups = "drop") |> 
  mutate(multicandidate = ifelse(
    candidates > 2, 1, 0
  )) |> 
  left_join(
    ballots |> 
      filter(date > "2013-01-01" & 
               result %in% c("won", "won_uncontested")) |> 
      select(date, seat, winner = coalition), 
    by = c("date", "seat")
  ) |> 
  filter(candidates != 1) |> 
  mutate(count = 1) |> 
  ggplot(aes(x = candidates, y = count)) + 
  geom_col(width = 1, 
                 aes(fill = winner)) +
  facet_wrap(~ federal, scales = "free") + 
  scale_fill_viridis_d(option = "magma") + 
  scale_x_continuous(breaks = seq(0, 7, 1)) + 
  theme(strip.background = element_rect(fill = "grey20")) + 
  labs(title = "Three-cornered fights are usually won by PH or BN", 
       subtitle = "Only electoral races post UMNO collapse (2018); only counts candidates which won more than 5% of the vote share", 
       y = "Number of electoral races", 
       n = "Number of candidates", 
       fill = "Winner")
  
  

```

<br>

The primary beneficiaries of three-cornered fights after the collapse of UMNO have been Pakatan Harapan and BN. 

<br>

Only a minority of PH/PR seats were won with less than 50% of the votes 

```{r}
ballots |> 
  filter(result %in% c("won", "won_uncontested") & 
           coalition %in% c("PH", "PR")) |> 
  mutate(year = year(date)) |> 
  ggplot(aes(x = votes_perc / 100, group = year, fill = -year)) + 
  geom_histogram() + 
  scale_x_continuous(labels = percent) + 
  facet_wrap(~ federal)
  
```
<br>

The margin by which PH/PR has won has been trending downwards

```{r}
ballots |> 
  filter(coalition %in% c("PH", "PR")) |> 
  mutate(year = year(date)) |> 
  ggplot(aes(x = str_sub(year, 3L), y = votes_perc / 100)) + 
  geom_jitter(aes(colour = result)) + 
  scale_colour_viridis_d(option = "plasma") + 
  scale_y_continuous(labels = percent, 
                     breaks = seq(0, 1, .1)) +
  facet_wrap(~ federal) + 
  theme(strip.background = element_rect(fill = "grey20")) + 
  labs(x = "Year", 
       y = "PH/PR vote share")

```

<br>

The footprint of PH is narrowing. 

Three-cornered fights are more commonplace 

PH/PR almost never wins when the percentage of votes it garners is less than 40%. 

Anwar is compromised by his sodomy conviction, so PH is not able to effectively enter into the Islamisation race. Also, is there a moderate Islamic position that the DAP can accept? 



```{r}
census_fed |> 
  filter(population_density < 600 & pc_bumi < .75)
```


```{r}
ballots |> 
  filter(election == "GE-15" & coalition == "PH" & result %in% c("won", "won_uncontested")) 
```



```{r}
multiracial |> 
  filter(multiracial == 1) |> 
  left_join(
    ballots |> 
      filter(result %in% c("won", "won_uncontested")) |> 
      select(seat, date, winner_ethnicity = ethnicity), 
    by = c("seat", "date")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |> 
  select(-c(malay_only:other_only)) |> 
  pivot_longer(
    cols = malay_chinese:sabah_sarawak_other, 
    names_to = "combination", 
    values_to = "value"
  ) |> 
  filter(value > 0) |> 
  mutate(
    combination = ifelse(
      combination %in% c(
        "malay_chinese_other", 
        "malay_sarawak", 
        "indian_other", 
        "sabah_other", 
        "sarawak_other", 
        "malay_chinese_other", 
        "malay_chinese_sabah", 
        "malay_chinese_sarawak"
      ), 
      "other_combination", 
      combination
    )) |> 
  group_by(combination) |> 
  mutate(sort = sum(value)) |> 
  ungroup() |> 
  mutate(combination = str_replace_all(combination, "_", "/"), 
         combination = str_to_title(combination)) |> 
  ggplot(aes(
    x = value, 
    y = fct_reorder(combination, sort), 
    fill = winner_ethnicity
  )) + 
  geom_col() +
  scale_fill_viridis_d() +
  labs(title = "Multiethnic electoral races and their results",
       x = "Number of electoral races", 
       y = "", 
       fill = "Winner ethnicity") +
  facet_wrap(~ federal, scales = "free_x") + 
  theme(strip.background = element_rect(fill = "grey20"))
```



```{r fig.height=7}
multiracial |> 
  filter(multiracial == 1) |> 
  left_join(
    ballots |> 
      filter(result %in% c("won", "won_uncontested")) |> 
      select(seat, date, winner_ethnicity = ethnicity), 
    by = c("seat", "date")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |> 
  select(-c(malay_only:other_only)) |> 
  pivot_longer(
    cols = malay_chinese:sabah_sarawak_other, 
    names_to = "combination", 
    values_to = "value"
  ) |> 
  mutate(winner_ethnicity = ifelse(
    winner_ethnicity == "Orang Asli", "Other", winner_ethnicity
  )) |> 
  ggplot(aes(x = decade, 
             y = value, 
             fill = winner_ethnicity)) + 
  geom_col() + 
  scale_fill_viridis_d() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) + 
  facet_wrap(~ federal, scales = "free_y") + 
  labs(
    title = "Results of multiethnic electoral races"
  )

```

```{r}
multiracial |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |>
  group_by(decade, federal) |> 
  summarise(n = n(), 
            multiethnic = sum(multiracial), 
            .groups = "drop")  |> 
  mutate(single_race = n - multiethnic,
         multiethnic_pc = multiethnic/ n, 
         single_race_pc = single_race/ n) |> 
  select(decade, federal, n, multiethnic_pc, single_race_pc) |> 
  pivot_longer(cols = multiethnic_pc:single_race_pc) |>
  ggplot(aes(x = decade, y = value, fill = name)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  facet_wrap(~ federal, scales = "free_y") +
  scale_fill_viridis_d(option = "cividis", begin = .2, end = .9)
  
  glimpse()
```

Instead of east malaysia, maybe by multiethnic fight?

Let's take a look at Chinese candidates, the largest minority group

non-Malay representation?

```{r}
ballots |> glimpse()
```

```{r}
ballots |> 
  mutate(votes_perc = ifelse(result == "won_uncontested", 100, votes_perc)) |> 
  filter(votes_perc > 3 & ethnicity == "Chinese") |> 
  mutate(count = 1) |> 
  group_by(party, federal) |> 
  summarise(count = sum(count)) |> 
  arrange(desc(count))
```

```{r}
ballots |> 
  mutate(count = 1) |> 
  
```




```{r}
candidates |> 
  mutate(east_malaysia = 
           ifelse(str_detect(state, 
                             "Sabah|Sarawak|Labuan"), 
                  "Yes", "No")) |> 
  group_by(parlimen, east_malaysia, election) |> 
  summarise(
    total_votes = sum(votes), 
    winning_votes = sum(votes[result == 1]), 
    losing_votes = sum(votes[result != 1]), 
    .groups = "drop"
  ) |> 
  mutate(margin = (winning_votes - losing_votes) / total_votes) |> 
  ggplot(aes(x = margin, fill = fct_rev(east_malaysia), group = east_malaysia)) + 
  geom_histogram() + 
  scale_x_continuous(labels = percent) + 
  scale_fill_manual(values = c("#90e0ef", "#03045e")) + 
  labs(x = "Margin of victory", 
       y = "Number of electoral races", 
       title = "Most electoral races in Malaysia are competitive", 
       subtitle = "Margins below 0% are three-cornerned fights", 
       fill = "East Malaysia")

```


Map of where multiracial races are

```{r}
multiracial |>
  filter(federal == "Federal") |> 
  mutate(year = year(date),
         decade = year - year %% 10,
         count = 1) |> 
  group_by(decade) |> 
  summarise(count = sum(count), 
            multiracial = sum(multiracial), 
            .groups = "drop") |> 
  mutate(pc = multiracial / count) |> 
  ggplot(aes(x = decade, y = pc)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) + 
  scale_y_continuous(labels = percent) + 
  labs(x = "", 
       y = "% of multiracial election races", 
       title = "% of Federal election races")
  
```

```{r}
multiracial |> glimpse()
```
```{r}
ballots |> count(election)

multiracial |> glimpse()
  
```


```{r fig.height=8}
census_fed |> 
  left_join(
    multiracial |> 
      filter(election %in% c("GE-12", "GE-13", "GE-14", "GE-15")) |> 
      mutate(type = case_when(
        multiracial == 1 ~ "Multiethnic", 
        malay_only == 1 ~ "Malay only", 
        chinese_only == 1 ~ "Chinese only", 
        indian_only == 1 ~ "Indian only", 
        sabah_only == 1 ~ "Bumi Sabah only", 
        sarawak_only == 1 ~ "Bumi Sarawak only"
      )) |> 
      select(seat, type, election), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_bumi = population_total * pc_bumi) |> 
  ggplot(aes(x = population_total, 
             y = pc_bumi)) + 
  geom_point(aes(colour = type, 
                 size = population_bumi)) +
  scale_x_log10(labels = comma) + 
  scale_y_continuous(labels = percent) +
  scale_colour_viridis_d() +
  scale_size_continuous(labels = comma) +
  labs(x = "Population of constituency", 
       y = "% of population who are Bumiputera", 
       colour = "Ethnicity of\ncandidates", 
       size = "2020 Bumiputera\npopulation") +
  guides(colour = guide_legend(
    override.aes = list(size = 2)
  )) + 
  facet_wrap(~ election)
```


```{r}
census_fed |> 
  left_join(
    multiracial |> 
      filter(election == "GE-15") |> 
      mutate(type = case_when(
        multiracial == 1 ~ "Multiethnic", 
        malay_only == 1 ~ "Malay only", 
        chinese_only == 1 ~ "Chinese only", 
        indian_only == 1 ~ "Indian only", 
        sabah_only == 1 ~ "Bumi Sabah only", 
        sarawak_only == 1 ~ "Bumi Sarawak only"
      )) |> 
      select(seat, type), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_bumi = population_total * pc_bumi) |> 
  ggplot(aes(x = population_density, 
             y = pc_bumi)) + 
  geom_point(aes(colour = type, 
                 size = population_total)) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = percent) +
  scale_colour_viridis_d() +
  scale_size_continuous(labels = comma) +
  labs(x = "Population density", 
       y = "% of population who are Bumiputera", 
       colour = "Ethnicity of\ncandidates", 
       size = "Total\npopulation") +
  guides(colour = guide_legend(
    override.aes = list(size = 3)
  ))
```

<br>

No Malay candidates ran where the population of non-Malays were 

Maybe put a red circle around the constituencies which were particularly competitive

```{r}
ballots |> 
  filter(rank == 3 & votes_perc > .03 & federal == "Federal") 

ballots |> filter(votes_perc > .1 & federal == "Federal") |> 
  group_by(date, seat, ethnicity) |> 
  summarise(votes = sum(votes))
```



<br>

```{r}

census_fed |> 
  mutate_at(
    vars("pc_bumi", "pc_chinese", "pc_indian", "pc_other", 
         "poverty_incidence", "sex_male", "sex_female"),
    ~ .x * population_total
  ) |> 
  rename_all(
    .funs = ~ gsub("pc_", "", .)) |> 
  pivot_longer(
    cols = bumi:other, 
    names_to = "ethnicity", 
    values_to = "value"
  ) |> 
  mutate(ethnicity = str_to_title(ethnicity)) |> 
  select(parlimen, 
         ethnicity, 
         value, 
         population_total) |> 
  ggplot(aes(y = value, fill = ethnicity, 
             x = fct_reorder(str_sub(parlimen, start = 1L, end = 5L), 
                             population_total))) + 
  geom_col(width = 1) + 
  scale_fill_viridis_d(option = "plasma") + 
  theme(axis.text.x = element_text(size = .5, angle = 90)) +
  scale_y_continuous(labels = number_format(scale = .001, suffix = "k"), 
                     breaks = seq(0, 700000, 50000)) + 
  labs(x = "Parliamentary seat", 
       y = "Constituency population", 
       title = "Massive parliamentary misapportionment", 
       subtitle = "A vote in the smallest constituency (Kanowit) is is worth 27.8 votes in the largest constituency (Bangi)", 
       fill = "")

```

```{r}
ballots |> 
  filter(state %in% c("Sabah", "Sarawak")) |> 
  mutate(ethnicity = ifelse(
    ethnicity == "Orang Asli", "Other", ethnicity
  )) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1) |> 
  group_by(state, decade, ethnicity, federal) |> 
  summarise(count = sum(count)) |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, 
             y = pc, 
             fill = ethnicity)) + 
  geom_col(position = "stack") + 
  facet_wrap(~ federal + state) + 
  scale_y_continuous(labels = percent) + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  scale_fill_viridis_d() + 
  theme(axis.text.x = element_text(angle = 30, vjust = .5), 
        strip.background = element_rect(fill = "black")) + 
  labs(title = "Ethnicity of election candidates", 
       y = "% of candidates", 
       x = "") 
```






```{r}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |> 
  filter(rank <= 3) |> 
  mutate(party = fct_lump(party, n = 12)) |> 
  group_by(party, sex) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(party) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates, 
         sex = ifelse(sex == "F", "Female", "Male")) |> 
  ggplot(aes(x = pc, y = fct_rev(party))) + 
  geom_col(aes(fill = sex)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of candidates", 
       y = "Party", 
       title = "Only 10.5% of candidates since 2006 have been women", 
       subtitle = "Only showing politicians who ran since the redelineation in 2006", 
       fill = "Sex")
```

PKR has backtracked on its commitment to one-person-one-vote in its [internal elections](https://www.iseas.edu.sg/articles-commentaries/iseas-perspective/2024-21-a-deep-dive-into-malaysias-peoples-justice-party-pkr-by-james-chai/): executive committees at the national and division levels are now chosen by delegates. 

Efforts have been made in recent years to ease this burden. First, PKR adopted a hybrid selection system for the first time in 2022,[69] where the executive committees at the national and division levels are chosen by delegates (see Table 6 above). This halves the maximum votes available to the members, which had proven to be too many for members in the first place (‘over-democracy’).[70] Two, introducing online voting (via the ‘ADIL’ application) that will minimise logistics and potential cheating. In 2022, more




[James Chin](https://www.commonwealthroundtable.co.uk/commonwealth/eurasia/malaysia/research-article-anwars-long-walk-to-power-the-2022-malaysian-general-elections/) at the Commonwealth Roundtable argues that there is no middle ground left in Malaysian politics. 

```{r}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  count(party, sort = TRUE)
```
