---
title: "Malaysian Elections"
subtitle: "Part two: Ethnicity and Election Predictions"
author: "Sean Ng"
organization: "AIMdata"
date-modified: "27 April 2026"
execute: 
  echo: false
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 9)


library(tidyverse)
library(here)
library(janitor)
library(scales)
library(tidytext)
library(widyr)
library(ggraph)
library(patchwork)
library(kableExtra)
library(viridis)
library(stringr)
library(sf)
library(ggmagnify)
library(ggthemes)
library(glmnet)
library(caret)
library(plotly)

`%out%` <- Negate(`%in%`)
options(scipen = 100)
theme_set(theme_light())
range_wna <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

last_comma <- function(x, sep = ","){
  x <- sapply(strsplit(x, sep),"[")
  trimws(x[[length(x)]])
}
```




```{r data}

census_dun <- read_csv("./data/census_dun.csv") |> 
  mutate(population_bumi = ethnicity_proportion_bumi * population_total / 100, 
         population_chinese = ethnicity_proportion_chinese * population_total/ 100, 
         population_indian = ethnicity_proportion_indian * population_total / 100, 
         population_other = ethnicity_proportion_other * population_total / 100)


gini <- read_csv("./data/hh_inequality_parlimen.csv") |> 
  filter(date == "2024-01-01")


sf::sf_use_s2(FALSE)

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))

federal_territories <- tribble(
  ~parlimen, ~population_total, ~pc_bumi, ~pc_chinese, ~pc_indian, ~pc_other, ~sex_male, ~sex_female, ~population_density, ~income_avg, ~expenditure_avg, ~poverty_incidence, ~area_km2,
  "P.114 Kepong", 106199, .202, .729, .063, .006, .588, .412, 8850, 13087, 6500, .003, 12,
  "P.115 Batu", 219132, .566, .301, .129, .004, .601, .399, 10957, 11019, 5817, 0, 20,
  "P.116 Wangsa Maju", 215600, .617, .306, .07, .008, .537, .463, 13475, 11153, 5768, 0, 16,
  "P.117 Segambut", 253715, .324, .558, .107, .012, .512, .479, 4975, 20521, 9845, 0, 51,
  "P.118 Setiawangsa", 147095, .697, .214, .085, .004, .519, .481, 9193, 12281, 6534, .003, 16, 
  "P.119 Titiwangsa", 122096, .677, .233, .076, .014, .519, .481, 8140, 11921, 6823, .005, 15,
  "P.120 Bukit Bintang", 120529, .271, .565, .155, .009, .565, .435, 5739, 13677, 7401, 0, 21, 
  "P.121 Lembah Pantai", 148094, .591, .262, .138, .009, .509, .491, 7405, 13081, 6186, 0, 20,
  "P.122 Seputeh", 322511, .311, .573, .112, .004, .510, .490, 10404, 13489, 7764, .005, 31,
  "P.123 Cheras", 135823, .322, .596, .076, .007, .527, .473, 8489, 12031, 5979, .002, 16,
  "P.124 Bandar Tun Razak", 191318, .679, .234, .083, .004, .509, .491, 7653, 11266, 6002, 0, 25,
  "P.166 Labuan", 95120, .862, .117, .011, .01, .514, .486, 1034, 8319, 4097, .031, 96,
  "P.125 Putrajaya", 109202, .979, .012, .006, .003, .509, .491, 2229, 12840, 7980, .004, 49
  
)

census_fed <- census_dun |> 
  mutate(population_poor = poverty_incidence * population_total / 100) |> 
  group_by(parlimen) |> 
  summarise_at(
    c("population_total", 
      "population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ sum(.x, na.rm = TRUE)
  ) |> 
  mutate_at(
    c("population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ .x / population_total
  ) |>  
  rename(
    pc_bumi = population_bumi, 
    pc_chinese = population_chinese, 
    pc_indian = population_indian, 
    pc_other = population_other, 
    poverty_incidence = population_poor) |> 
  left_join(
    census_dun |>
      mutate(
        income = income_avg * household_total,
        expenditure = expenditure_avg * household_total) |>
      group_by(parlimen) |>
      summarise(
        household_total = sum(household_total),
        income = sum(income),
        expenditure = sum(expenditure)) |>
      mutate(
        income_avg = income / household_total,
        expenditure_avg = expenditure / household_total) |>
      select(parlimen,
             income_avg, 
             expenditure_avg), 
    by = "parlimen") |> 
  left_join(
    parlimen_geo |> 
      st_drop_geometry() |> 
      select(parlimen, area_km2), 
    by = "parlimen"
  ) |> 
  mutate(population_density = population_total / area_km2) |> 
  rbind(federal_territories) |> 
  left_join(
    census_dun |> 
      distinct(parlimen, state), 
    by = "parlimen"
  ) |>
  left_join(
    gini |> 
      select(parlimen, gini), 
    by = "parlimen"
  ) |> 
  mutate(state = case_when(
    is.na(state) & str_detect(parlimen, "Labuan") ~ "Labuan", 
    is.na(state) & str_detect(parlimen, "Putrajaya") ~ "Putrajaya", 
    is.na(state) ~ "Kuala Lumpur", 
    TRUE ~ state
  )) |> 
  mutate(east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                                1, 0))

ballots <- read_csv("./data/consol_ballots.csv") |> 
  mutate(
    age = ifelse(age == -1, NA_real_, age)
  ) |> 
  mutate(federal = ifelse(
    str_detect(seat, "P.0|P.1|P.2"), 
    "Federal", 
    "State"))

stats <- read_csv("./data/consol_stats.csv")

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))

multiracial_races <- ballots |> 
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(count = 1) |> 
  # I think this filter is fine
  filter(rank <= 3) |> 
  group_by(state, seat, date, federal, election) |> 
  summarise(malay_candidates = sum(count[ethnicity == "Malay"]), 
            chinese_candidates = sum(count[ethnicity == "Chinese"]), 
            indian_candidates = sum(count[ethnicity == "Indian"]), 
            sabah_candidates = sum(count[ethnicity == "Bumi Sabah"]), 
            sarawak_candidates = sum(count[ethnicity == "Bumi Sarawak"]), 
            other_candidates = sum(count[ethnicity == "Other"]), 
            .groups = "drop") |> 
  mutate(total_candidates = malay_candidates + chinese_candidates + indian_candidates + 
           sabah_candidates + sarawak_candidates + other_candidates) |> 
  pivot_longer(cols = malay_candidates:other_candidates, 
               names_to = "candidate_ethnicity", 
               values_to = "value") |> 
  mutate(candidate_ethnicity = str_remove_all(candidate_ethnicity, "_candidates")) |> 
  filter(value != 0) |>
  mutate(multiracial = ifelse(value != total_candidates, 1, 0)) |> 
  mutate(combination = ifelse(
    multiracial == 0, paste0(candidate_ethnicity, "_only"), "multiethnic"
  )) |> 
  distinct(state, seat, date, federal, election, multiracial, combination)
```

**Acronyms**
```{r table-acronyms, echo = FALSE, message=FALSE, warning=FALSE}
tribble(
  ~acronym, ~name, 
  "BN", "Barisan Nasional",  
  "DAP", "Democratic Action Party",
  "GPS", "Gabnungan Parti Sarawak",
  "GRS", "Gabungan Rakyat Sabah", 
  "GS", "Gagasan Sejahtera", 
  "MCA", "Malaysian Chinese Association", 
  "MIC", "Malaysian Indian Congress", 
  "PAS", "Parti Islam Se-Malaysia", 
  "PH", "Pakatan Harapan", 
  "PKR", "Parti Keadilan Rakyat", 
  "PN", "Perikatan Nasional", 
  "SUPP", "Sarawak United People's Party", 
  "UMNO", "United Malays National Organisation", 
  "USA", "United Sabah Alliance"
) |>  
  kable(col.names = NULL, format = "html") |> 
  kable_classic(bootstrap_options = c("condensed"), 
                full_width = FALSE, 
                position = "float_left")
```

<br>

Part one of this report may be read [here](https://aimdata-labs.github.io/malaysian_elections/). 

# Racialisation of elections

According to an [Architects of Diversity survey](https://www.aodmalaysia.org/sods), 64% of Malaysians reporting experiencing some for of discrimination in the past 12 months, with 38% saying the discrimination was related to socioeconomic status, 33% saying it was related to age and 32% saying it was related to ethnicity. 

Hindus were the most likely to have experienced religious-based discrimination (40%), compared to their Christian (26%), Buddhists (22%) and Muslims (20%). 59% of Muslim respondents believe their own religious group faced a lot discrimination, with the figure being 63% amongst Buddhists, 76% amongst Christians and 81% amongst Hindus. 

According to [Pusat KOMAS](https://komas.org/download/Malaysia-Racism-Report_A-Decade-In-Review-2015-2024.pdf), election years tend to see "sharp increases in racial discrimination narratives [...] Political parties have consistently used race and religion as mobilization tools". They add, "politics accounted for just over 43% of all recorded racial discrimination incidents between 2015 and 2024" and that "racial and religious politics remain the primary diver of racial discrimination in Malaysia". 

The next general election is likely to witness another surge in racial politics, particularly with PAS and Bersatu continuing to consolidate their Malay-Muslim voter bases. 

<br><br><br>

# Multiethnic electoral races

```{r map}

parlimen_geo |> 
  left_join(
    multiracial_races |>
      filter(federal == "Federal") |>
      select(seat, multiracial, combination) |>
      group_by(seat) |>
      slice_max(order_by = multiracial) |>
      ungroup() |>
      distinct(seat, multiracial, combination), 
    by = c("parlimen" = "seat") 
  ) |> 
  mutate(multiracial = ifelse(multiracial == 1, "Multiethnic", "Single ethnicity")) |> 
  ggplot() + 
  geom_sf(aes(fill = multiracial), 
          linewidth = .01, 
          colour = "grey70") + 
  scale_fill_viridis_d(option = "magma", 
                       begin = .2, 
                       end = .9) + 
  # Only one inset can work at one time
  # Solution proposed here https://github.com/hughjonesd/ggmagnify/issues/30 
  # Does not work
  geom_magnify(
    from = c(101.223422,  101.953542, 2.805661, 3.383667), 
    to = c(105, 108, 3.5, 5.9), 
    shadow = FALSE,
    proj = "single"
  ) + 
    #theme_map() + 
  labs(
    fill = "",
    title = "Federal constituencies which have ever had multiethnic electoral races"
  ) + 
  theme_void() +

parlimen_geo |> 
  left_join(
    census_fed |>
      select(parlimen, pc_bumi), 
    by = "parlimen"
  ) |> 
  ggplot() + 
  geom_sf(aes(fill = pc_bumi), 
          linewidth = .01, 
          colour = "grey70") + 
  scale_fill_viridis(option = "magma", 
                       begin = .2, 
                       end = .9, 
                     labels = percent, 
                     breaks = seq(.1, .9, .1)) + 
  # Only one inset can work at one time
  # Solution proposed here https://github.com/hughjonesd/ggmagnify/issues/30 
  # Does not work
  geom_magnify(
    from = c(101.223422,  101.953542, 2.805661, 3.383667), 
    to = c(105, 108, 3.5, 5.9), 
    shadow = FALSE,
    proj = "single"
  ) + 
    #theme_map() + 
  labs(
    fill = "",
    title = "Federal constituencies by the proportion of their population who are Bumiputera"
  ) + 
  theme_void() +
  
  plot_layout(ncol = 1)
  
```

<br>

43% (97 out of 222 constituencies) have ever had multiethnic electoral races. These constituencies align largely with the areas with the highest concentrations of non-Bumiputeras. 



<br>

```{r race-type}
ballots |>
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1) |>
  # This is just to isolate the number of electoral races
  filter(result %in% c("won", "won_uncontested")) |> 
  group_by(
    decade, federal, combination
  ) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count), 
         combination = case_when(
           combination == "sabah_only" ~ "bumi_sabah_only", 
           combination == "sarawak_only" ~ "bumi_sarawak_only", 
           TRUE ~ combination), 
         combination = str_to_title(str_replace_all(combination, "_", " "))
         ) |> 
  ungroup() |>
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, y = pc, fill = fct_relevel(combination, 
                                   c("Bumi Sabah Only", 
                                     "Bumi Sarawak Only", 
                                     "Chinese Only", 
                                     "Indian Only", 
                                     "Malay Only", 
                                     "Other Only", 
                                     "Multiethnic")))) + 
  geom_col() + 
  facet_wrap(~ federal) + 
  scale_y_continuous(labels = percent) +
  scale_fill_manual(
    values = c(
      "Bumi Sabah Only" = "#440154FF", 
      "Bumi Sarawak Only" = "#414487FF",
      "Chinese Only" = "#2A788EFF",
      "Indian Only" = "#22A884FF",
      "Malay Only" = "#7AD151FF",
      "Other Only" = "#FDE725FF",
      "Multiethnic" = "#FCFDBFFF"
    )
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(fill = "",
       title = "Multiethnic electoral races were least common during the 90s and 00s", 
       x = "", 
       y = "") + 
  theme(strip.background = element_rect(fill = "grey20"), 
        axis.text.x = element_text(size = 7))
```
<br>

Most constituencies in Malaysia are multi-ethnic in nature: `r census_fed |> filter(pc_bumi > .75) |> nrow()` out of the `r census_fed |> nrow()` federal constituencies have at least 25% of their populations being racial minorities. The latest redelineation was done in 2006 and the plots in this section will focus on the years since then. 


<br> 

```{r multiethnic-race}
ballots |>
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1, 
         ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |>
  filter(result %in% c("won", "won_uncontested") & combination == "multiethnic") |> 
  group_by(
    decade, federal, combination, ethnicity
  ) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count)) |> 
  ungroup() |>
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, y = pc, fill = fct_relevel(ethnicity, 
                                   c("Bumi Sabah", 
                                     "Bumi Sarawak", 
                                     "Chinese", 
                                     "Indian", 
                                     "Malay", 
                                     "Other")))) + 
  geom_col() + 
  facet_wrap(~ federal) + 
  scale_y_continuous(labels = percent) +
  scale_fill_manual(
    values = c(
      "Bumi Sabah" = "#440154FF", 
      "Bumi Sarawak" = "#414487FF",
      "Chinese" = "#2A788EFF",
      "Indian" = "#22A884FF",
      "Malay" = "#7AD151FF",
      "Other" = "#FDE725FF")
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(fill = "Combination",
       title = "Multiethnic electoral races were most commonly won by Chinese candidates",
       subtitle = "Though this has been on the downtrend since the 00s",
       x = "", 
       y = "") + 
  theme(strip.background = element_rect(fill = "grey20"), 
        axis.text.x = element_text(size = 7))
```


<br>

Commonsensically, given the racialised nature of Malaysian politics, Bumiputera candidates are far more likely to win in areas that have high proportions of Bumiputera candidates. Non-bumiputera candidates almost never win seats in constituencies with more than 75% Bumiputera residents and Bumiputera candidates tend only to win when a constituency has a population that is more than 50% Bumi. 

<br>


```{r winner-bumi}
ballots |>
  filter(date > "2006-01-01") |> 
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  left_join(
    census_fed |> 
      select(parlimen, pc_bumi), 
    by = c("seat" = "parlimen")
  ) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  mutate(
    bumiputera = ifelse(ethnicity %in% c("Malay", "Bumi Sabah", "Bumi Sarawak"),
                        "Bumiputera winner",
                        "non-Bumi winner"),
    east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                           "East Malaysia", "Peninsula")) |> 
  ggplot(aes(x = pc_bumi, fill = east_malaysia)) + 
  geom_histogram() + 
  scale_fill_viridis_d(begin = .1, end = .9) +
  scale_x_continuous(labels = percent) +
  facet_wrap(~ bumiputera) + 
  labs(
    title = "Electoral races by % of the constituency who are Bumiputera", 
    subtitle = "Shows races since the 2006 redelineation", 
    fill = "", 
    y = "Number of electoral races", 
    x = "% of constituency who are Bumiputera"
  ) + 
  theme(strip.background = element_rect(fill = "grey20"))

```

<br>

The coalition which benefited primarily from this racialised dimension to elections has been Barisan Nasional. 

Almost all of BN's electoral wins come from constituencies which are more than 75% Bumiputera. This is the same for Perikatan Nasional. Conversely, constituencies which are more than 75% Bumiputera almost never vote for parties under the Pakatan Harapan coalition (or its predecessor Pakatan Rakyat). 

<br>

```{r coalition-bumi}
ballots |>
  filter(date > "2006-01-01") |> 
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  left_join(
    census_fed |> 
      select(parlimen, pc_bumi), 
    by = c("seat" = "parlimen")
  ) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  mutate(
    bumiputera = ifelse(ethnicity %in% c("Malay", "Bumi Sabah", "Bumi Sarawak"),
                        "Bumiputera winner",
                        "non-Bumi winner"),
    east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                           "East Malaysia", "Peninsula")) |> 
  ggplot(aes(x = pc_bumi, fill = coalition)) + 
  geom_histogram() + 
  scale_fill_manual(values = c(
    "ALONE" = "grey50", 
    "BN" = "#1D1147FF", 
    "GPS" = "#51127CFF", 
    "GRS" = "#822681FF", 
    "GS" = "#B63679FF", 
    "PH" = "#E65164FF", 
    "PN" = "#FB8861FF", 
    "PR" = "#FEC287FF", 
    "USA" = "#FCFDBFFF"
  )) +
  scale_x_continuous(labels = percent) +
  facet_wrap(~ bumiputera) + 
  labs(
    title = "Electoral races by % of the constituency who are Bumiputera", 
    subtitle = "Shows races since the 2006 redelineation", 
    fill = "Coalition", 
    y = "Number of electoral races", 
    x = "% of constituency who are Bumiputera"
  ) + 
  theme(strip.background = element_rect(fill = "grey20"))
```

<br> 


A significant minority of non-Bumiputera MPs were from BN. Additionally, most of those wins came from constituencies with higher Bumiputera populations, indicating that party loyalty (which itself has racial connotations) was, at one point, stronger than the ethnicity of individual candidates. 

<br><br><br>

# Ethnic makeup of parties and coalitions

Amongst the various coalitions, East Malysian Coalitions and Gagasan Sejahtera were the most racially homogenous. Proportionally, Pakatan Harapan and Pakatan Rakyat both fielded more Malay candidates than Barisan Nasional. 

However, as noted by [Azzubair, Kartini and Nazri](https://www.frontiersin.org/journals/political-science/articles/10.3389/fpos.2026.1721966/full), whilst BN might be multiethnic in nature, the coalition was not consociational, but was "sustained more by UMNO’s control and patronage than by genuine power-sharing". 


<br>

```{r coalition-ethncity}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |> 
  filter(rank <= 3) |>
  group_by(coalition, ethnicity) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(coalition) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates) |> 
  ggplot(aes(x = pc, y = fct_rev(coalition))) + 
  geom_col(aes(fill = ethnicity)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of candidates", 
       y = "Coalition", 
       title = "Candidates fielded by ethnicity", 
       subtitle = "Only showing politicians who ran since the redelineation in 2006", 
       fill = "Ethnicity")
  
  
```

<br>

The plot below shows the breakdown of candidates by ethnicity and party. No party reflects the racial make-up of the country. Amanah, Bersatu, PAS and UMNO are Malay parties; DAP, Gerakan and the MCA are Chinese parties. The MIC is an Indian party, PBB is a Sabahan party and WARISAN is a Sarawakian party. 

When we look at party level, we note that BEBAS and PKR are the only true multiracial parties, with all other parties being dominated by a main ethnic group. 

<br> 


```{r ethnicity-party-candidates}
ballots |> 
  filter(date > "2006-01-01") |> 
  mutate(count = 1) |>
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(party = fct_lump(party, n = 12)) |> 
  group_by(party, ethnicity) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(party) |> 
  mutate(total_candidates = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total_candidates) |> 
  ggplot(aes(x = pc, y = fct_rev(fct_relevel(party, 
                                             c("Other", "UMNO", "PAS", "PKR", 
                                               "BEBAS", "DAP", "MCA", "BERSATU", 
                                               "AMANAH", "WARISAN", "PBB","GERAKAN", "STAR"
                                             ))))) + 
  geom_col(aes(fill = ethnicity)) + 
  geom_vline(xintercept = .25, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .5, linetype = "longdash", colour = "white", alpha = .4) +
  geom_vline(xintercept = .75, linetype = "longdash", colour = "white", alpha = .4) +
  scale_fill_viridis_d() + 
  scale_x_continuous(label = percent) + 
  labs(x = "% of candidates", 
       y = "Party", 
       title = "The only true multi-ethnic parties are BEBAS and PKR", 
       subtitle = "Only candidates since 2006; parties ordered by number of candidates", 
       fill = "Candidate\nethnicity")
  
  
```


<br><br><br>

# The collapse of UMNO and Islamic politics

According to [Deivasagayam](https://msocialsciences.com/index.php/mjssh/article/view/3248/2134), the two main cleavages in Malaysian society -- race and religion -- have converged, contributing to UMNO's downfall. He continues, "UMNO consistently positioned itself as the exclusive advocate and guardian of Malay rights and privileges". However, by engaging in an "Islamisation Race" to stave off PAS, UMNO inadvertently legitimised PAS in the eyes of Muslim voters, and when it advocated for Malay Islamic Supremacy, it set that stage for the convergence for race and religion which would later ultimately benefit PAS and BERSATU, which demonstrated the political viability of a Malay-Muslim governance structure which excluded minorities. 


We see, from the plot below, the extent of UMNO's electoral collapse. From its peak of `r filter(ballots, election == "GE-11" & result %in% c("won", "won_uncontested") & party == "UMNO") |> nrow()` seats under Badawi, it now holds less seats than either PAS, DAP, PKR or Bersatu. 

<br>


```{r parties-mp-share}
ballots |> 
  mutate(count = 1) |> 
  filter(result %in% c("won", "won_uncontested")) |> 
  filter(str_detect(election, "GE")) |> 
  mutate(year = year(date)) |> 
  group_by(election) |> 
  mutate(year = min(year)) |> 
  ungroup() |> 
  mutate(party = fct_lump(party, n = 12, w = votes)) |> 
  group_by(year, party) |> 
  summarise(mps = sum(count), 
            .groups = "drop") |> 
  ggplot(aes(x = year, y = mps, group = party)) + 
  geom_line(aes(colour = party), 
            linewidth = 1.05, 
            alpha = .8) + 
  # scale_colour_manual(
  #   values = c(
  #     "UMNO" = "#30123BFF",
  #     "MCA" = "#4454C4FF",
  #     "MIC" = "#4490FEFF",
  #     "BERSATU" = "#1FC8DEFF",
  #     "PAS" = "#29EFA2FF",
  #     "GERAKAN" = "#7DFF56FF",
  #     "PBB" = "#C1F334FF",
  #     "PBS" = "#F1CA3AFF",
  #     "USNO" = "#FE922AFF",
  #     "SUPP" = "#EA4F0DFF",
  #     "PKR" = "#BE2102FF",
  #     "DAP" = "#7A0403FF",
  #     "Other" = "grey50"
  #     
  #   )
  # ) +
  scale_colour_viridis_d(option = "turbo") +
  scale_x_continuous(breaks = seq(1955, 2026, 5)) +
  scale_y_continuous(breaks = seq(0, 120, 10)) +
  guides(colour = guide_legend(override.aes = list(linewidth = 2, 
                                                   alpha = 1))) + 
  labs(x = "", 
       y = "Number of MPs", 
       title = "The collapse of UMNO has left PAS and DAP as the largest parties", 
       subtitle = "Results from federal elections", 
       colour = "") + 
  theme(axis.text.x = element_text(size = 7))
```
<br>

However, when we examine their share of votes, we see that their parliamentary dominance was always more fragile than was perceived. At no time after 1960 did UMNO ever win more than 40% of the votes. Additionally, the Badawi peak of `r filter(ballots, election == "GE-11" & result %in% c("won", "won_uncontested") & party == "UMNO") |> nrow()` seats was accomplished with only 35% of the national vote, highlighting the unfairness of a first-past-the-post system. Though now, both UMNO and PKR are penalised by the first-past-the-post system, whereas PAS, DAP and Bersatu are each rewarded by it. 


Finally, though their share of votes is lower than it has ever been, UMNO still, marginally, received more votes than any other party in the most recent general election. 

<br> 

```{r parties-vote-share}
ballots |>  
  filter(str_detect(election, "GE")) |>
  mutate(year = year(date)) |> 
  group_by(election) |> 
  mutate(year = min(year)) |>
  ungroup() |> 
  mutate(party = fct_lump(party, n = 12, w = votes)) |>
  group_by(year, party) |> 
  summarise(votes = sum(votes), .groups = "drop") |>
  group_by(year) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = year, y = votes_pc, group = party)) + 
  geom_line(aes(colour = party), 
            linewidth = 1.05, 
            alpha = .8) + 
  scale_colour_viridis_d(option = "turbo") +
  scale_x_continuous(breaks = seq(1955, 2026, 5)) +
  guides(colour = guide_legend(override.aes = list(linewidth = 2, 
                                                   alpha = 1))) + 
  scale_y_continuous(labels = percent, breaks = seq(0, .6, .1)) +
  labs(x = "", 
       y = "Share of popular vote", 
       title = "UMNO still has, very marginally, the largest share of votes",
       subtitle = "But it has never won more than 40% of votes since the formation of the Federation", 
       colour = "") + 
  theme(axis.text.x = element_text(size = 7))
  
  
```

<br> 

Recalling Deivasagayam's arguments, [Syaza](https://www.iseas.edu.sg/wp-content/uploads/2024/04/TRS12_24.pdf) concludes that "the sense of Islamic identity and belonging often supersedes the allure of economic development for Malay youths when choosing how to cast their vote". 

This speaks to not only to the Islamisation race that UMNO, PAS and Bersatu have engaged in -- 65% of respondents in Muslim Youth Survey saying that Muslims should only vote for Muslim leaders -- but also to an unfair economic system where whatever economic boons offered are too meagre and piecemeal, and insufficient to attract voters.

Pakatan Harapan seems unwilling to engage in the Islamisation Race or put forward a competing vision of a more moderate Islam, possibly -- I hypothesise -- out of fears that doing so would alienate its core non-Muslim base and anger conservative Muslims.




<br><br><br>

# Fragile coalitions

Whilst public support for BN has faded considerably, between BN and PN, there was majority support for a Malay-supremacist (*Ketuanan Melayu*) government, picking up more than 50% of votes in GE-15, an increase in the vote share garnered by BN and GS in GE-14. 

The current unity (*Madani*, PH-BN-GPS-GRS) government is heavily reliant on BN for both a parliamentary majority as well as a majority of the vote share. 

<br>

```{r general-elections-vote-share}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA"), "Other", coalition
  )) |> 
  filter(election %in% c("GE-15", "GE-14", "GE-13", "GE-12")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            candidates = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = votes_pc, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(labels = percent) + 
  labs(x = "% of votes", 
       y = "", 
       fill = "Coalition", 
       title = "Vote share by coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br>

The two largest coalitions in parliament are now PH and PN. But since neither were able to win a commanding mandate in GE-15 (or able to compromise with each other), BN was left to play kingmaker. 

<br>

```{r general-elections-seats}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA"), "Other", coalition
  )) |> 
  filter(election %in% c("GE-15", "GE-14", "GE-13", "GE-12") & 
           result %in% c("won", "won_uncontested")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            seats = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = seats, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(breaks = seq(0, 150, 20)) +
  labs(x = "Parliament seats", 
       y = "", 
       fill = "Coalition", 
       title = "Parliament seats by coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br>

The plots below show vote share amongst the major coalitions in state elections. We see that that PN has managed to garner a lead over PH in state polls. 

Recalling our earlier criticism of Anwar in [part one](https://aimdata-labs.github.io/malaysian_elections/), is his bid with GRS to increase the number of seats in Sabah meant to engineer a more permanent parliamentary majority without needing to win the popular vote? We emphasise that any further malapportionment would be, ultimately, an anti-democratic exercise. 

<br>

```{r state-elections-coalitions}
ballots |> 
  mutate(count = 1) |> 
  mutate(coalition = ifelse(
    coalition %in% c("USA", "ALONE", "GTA", "WARISAN-PLUS", "GASAK"), "Other", coalition
  )) |> 
  filter(election %in% c("SE-15", "SE-14", "SE-13", "SE-12")) |> 
  group_by(election, coalition) |>
  summarise(votes = sum(votes), 
            candidates = sum(count), 
            .groups = "drop") |> 
  group_by(election) |> 
  mutate(total_votes = sum(votes)) |> 
  ungroup() |> 
  mutate(votes_pc = votes / total_votes) |> 
  ggplot(aes(x = votes_pc, y = fct_rev(coalition), fill = coalition)) + 
  geom_col() + 
  facet_wrap(~ election, scales = "free_y") + 
  scale_fill_viridis_d(option = "magma", begin = .1, end = .9) + 
  scale_x_continuous(labels = percent) + 
  labs(x = "% of votes", 
       y = "", 
       fill = "Coalition", 
       title = "State election vote share by coalition") + 
  theme(strip.background = element_rect(fill = "grey30"))
```

<br> 

There is a stark racial component in Malaysian politics. [Syaza](https://www.iseas.edu.sg/wp-content/uploads/2024/04/TRS12_24.pdf) argues that there is a strong dichotomy between Malay and non-Malay voters' perceptions of the government. 85% of Malay voters felt that the Perikatan Nasional government protects the interest of the Malays, while 77% of non-Malay voters believed the government does not treat all races equally. This is key to understanding political choices among different ethnic groups in Malaysia, not only in the 15th General Election, but also in previous elections.

Given this strong racial component to elections, we will next use demographic data to predict Pakatan Harapan's performance in GE-16. 


<br><br><br>


# Predictions for the GE-16


```{r training-and-testing}

training <- census_fed |> 
  mutate(pc_bumi_peninsula = ifelse(
    east_malaysia == 0, pc_bumi, 0), 
    pc_bumi_east_malaysia = ifelse(
      east_malaysia == 1, pc_bumi, 0
    )) |> 
  select(
    parlimen, population_density, pc_bumi_peninsula, pc_bumi_east_malaysia, 
    pc_chinese, pc_indian, gini, income_avg
  ) |> 
  left_join(
    ballots |>
      filter(election == "SE-15") |>
      rbind(ballots |>
              filter(date == "2021-12-18" & state == "Sarawak")) |>
      left_join(census_dun |>
                  distinct(dun, parlimen), by = c("seat" = "dun")) |>
      group_by(parlimen) |>
      summarise(
        ph_votes = sum(votes[coalition == "PH"], na.rm = TRUE),
        total_votes = sum(votes, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(pc_ph_votes = ph_votes / total_votes), 
    by = "parlimen"
  ) |> 
  filter(!is.na(pc_ph_votes)) |>
  mutate(population_density = log(population_density)) |> 
  mutate_at(
    c("population_density", "pc_bumi_peninsula", "pc_bumi_east_malaysia", 
      "pc_chinese", "pc_indian", "gini", "income_avg"), 
    function(x){
      (x - mean(x) / sd(x))
    }
  ) |> 
  na.omit()

testing <- census_fed |> 
  mutate(pc_bumi_peninsula = ifelse(
    east_malaysia == 0, pc_bumi, 0), 
    pc_bumi_east_malaysia = ifelse(
      east_malaysia == 1, pc_bumi, 0
    )) |> 
  select(
    parlimen, population_density, pc_bumi_peninsula, pc_bumi_east_malaysia, 
    pc_chinese, pc_indian, gini, income_avg
  ) |> 
  left_join(
    ballots |> 
      filter(election == "GE-15") |> 
      group_by(seat) |> 
      summarise(ph_votes = sum(votes[coalition == "PH"], na.rm = TRUE), 
                total_votes = sum(votes, na.rm = TRUE), 
                .groups = "drop") |>
      mutate(pc_ph_votes = ph_votes / total_votes), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_density = log(population_density)) |> 
  mutate_at(
    c("population_density", "pc_bumi_peninsula", "pc_bumi_east_malaysia", 
      "pc_chinese", "pc_indian", "gini", "income_avg"), 
    function(x){
      (x - mean(x) / sd(x))
    }
  ) |> 
  na.omit()
```


```{r stepwise-model}
#| echo: false
#| output: false

stepwise_model <- step(lm(
  pc_ph_votes ~ population_density + pc_bumi_peninsula + pc_bumi_east_malaysia + 
  pc_chinese + pc_indian + gini + income_avg, 
  data = training, 
  direction = "backward"
))

# The stepwise model indicates that the most relevant variables to predicting 
# PH performance are log(population density), % of population who are peninsula bumis, 
# % of the population who are east malaysia bumis and % of the population who are indian

```


```{r glmnet-model}
glmnet_model <- train(
  pc_ph_votes ~ population_density + pc_bumi_peninsula + pc_bumi_east_malaysia + 
  pc_chinese + pc_indian + gini + income_avg, 
  data = training, 
  method = "glmnet", 
  metric = "RMSE", 
  trControl = trainControl(method = "LOOCV"), 
  preProcess = c("center", "scale"), 
  tuneLength = 10
)

```

```{r preds}
preds <- testing %>%
  mutate(
    aic_pred = predict(object = stepwise_model, newdata = .),
    glmnet_pred = predict(object = glmnet_model, newdata = testing)
  ) %>%
  mutate(pred = (aic_pred + glmnet_pred) / 2) |> 
  select(proj_ph_vote = pred, 
         step_pred = aic_pred, 
         glmnet_pred = glmnet_pred)

testing <- testing |> 
  cbind(preds)

```

In this section, we will try and predict Pakatan Harapan's performance (in this case, the percentage of votes we anticipate them to receive) in the upcoming general election. We are looking at PH's performance because it is the easiest to predict from past election performance and census data: BN's ongoing collapse adds elements that are difficult to account for with demographic data, as does the overlap between BN and PN voters.


```{r ph-win-chance, eval=FALSE}
ballots |> 
  filter(coalition == "PH") |>
  filter(votes_perc < 50) |> 
  mutate(ph_win = ifelse(result %in% c("won", "won_uncontested"), 1, 0)) %>% 
  lm(ph_win ~ votes_perc, data = .) |> 
  summary() |> 
  tidy()
```




```{r}
safe_seats <- testing |> 
  left_join(
    ballots |> 
      filter(election == "GE-15" & result %in% c("won", "won_uncontested")) |> 
      select(state, seat, coalition), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(
    ph_win_chance = ifelse(proj_ph_vote > .5, 1, proj_ph_vote * 0.01479184 * 100)) |> 
  filter(ph_win_chance > .5) |> 
  nrow()
```


Below, the performance of PH is GE-15 is plotted against their predicted performance in GE-16, with each point being a federal electoral constituency. First of, we note that the combined stepwise-glmnet model predicts that PH's vote share will shrink in GE-16. 

We note that the number of "safe" seats that PH can expect to win is only **`r safe_seats`** out of the 222 total seats in parliament. 


<br>


```{r plotting-results}
#| out-width: 100%
#| out-height: 100%
pred_plot <- testing |> 
  left_join(
    ballots |> 
      filter(election == "GE-15" & result %in% c("won", "won_uncontested")) |> 
      select(state, seat, coalition), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(
    ph_win_chance = ifelse(proj_ph_vote > .5, 1, proj_ph_vote * 0.01479184 * 100)) |> 
  ggplot(aes(x = pc_ph_votes, y = proj_ph_vote)) + 
  geom_point(aes(colour = ph_win_chance,
                 text = paste0(parlimen, ",", "\n", 
                               state, ",", "\n", 
                               "PH win chance: ", round(ph_win_chance * 100, 2), "%", "\n",
                               "Projected PH%: ", round(proj_ph_vote * 100, 2), "%", "\n", 
                               "GE-15 PH%: ", round(pc_ph_votes * 100, 2), "%", "\n", 
                               "GE-15 winner: ", coalition))) +
  geom_smooth(method = "lm") + 
  scale_colour_viridis(option = "plasma", 
                       labels = percent) + 
  scale_x_continuous(labels = percent, breaks = seq(0, 1, .1)) + 
  scale_y_continuous(labels = percent, breaks = seq(0, 1, .1)) +
  labs(colour = "Chance of\nPH win", 
       x = "% PH votes in GE-15", 
       y = "Projected % of PH votes in GE-16", 
       title = "Interactive plot of GE-15 results and predictions for GE-16", 
       subtitle = "Mouse over for details; drag and click to select and zoom; double-click legend select/deselect")

ggplotly(pred_plot, tooltip = c("text")) |> 
  layout(height = 500, width = 800, 
         margin = list(t = 50)) |> 
  config(displayModeBar = FALSE)
```


<br>

These predictions also underscore the instability of the Madani government's coalition. The results have been re-plotted below to show the GE-15 winner of each constituency. Marginal seats where PH may seak to expand its reach will largely only come at the expense of BN and vice versa. 

There are a fair number of marginal seats where PH received less than 40% of the votes in GE-15, but were still won by PH (many of these were three-cornered fights). These are all vulnerable for pickup by BN or PN. 

PH may extend their reach by running better races and fielding better candidates in Bagan Datuk in Perak, Titiwangsa in KL, Tampin and Jempol in Negeri Sembilan, Kuala Langat in Selangor and Padang Serai in Kedah. 

<br>

```{r results-coalition}
#| out-width: 100%
#| out-height: 100%
pred_plot_coalition <- testing |> 
  left_join(
    ballots |> 
      filter(election == "GE-15" & result %in% c("won", "won_uncontested")) |> 
      select(state, seat, coalition), 
    by = c("parlimen" = "seat")
  ) |> 
  ggplot(aes(x = pc_ph_votes, y = proj_ph_vote)) + 
  geom_point(aes(colour = coalition,
                 text = paste0(parlimen, ",", "\n", 
                               state, ",", "\n", 
                               "Projected PH%: ", round(proj_ph_vote * 100, 2), "%", "\n", 
                               "GE-15 PH%: ", round(pc_ph_votes * 100, 2), "%", "\n", 
                               "GE-15 winner: ", coalition))) +
  geom_smooth(method = "lm") + 
  scale_colour_viridis_d(option = "turbo") + 
  scale_x_continuous(labels = percent, breaks = seq(0, 1, .1)) + 
  scale_y_continuous(labels = percent, breaks = seq(0, 1, .1)) +
  labs(colour = "Winning\ncoalition\nin GE15", 
       x = "% PH votes in GE-15", 
       y = "Projected % of PH votes in GE-16", 
       title = "Interactive plot of GE-15 results and predictions for GE-16", 
       subtitle = "Mouse over for details; drag and click to select and zoom; double-click legend select/deselect")

ggplotly(pred_plot_coalition, tooltip = c("text")) |> 
  layout(height = 500, width = 800, 
         margin = list(t = 50)) |> 
  config(displayModeBar = FALSE)
```


<br><br><br>


## Three-cornered fights and shrinking vote share

Given the loose alliances between the various coalitions, three-cornered fights have become much more prevalent. 

The primary beneficiaries of three-cornered fights after the collapse of UMNO have been Pakatan Harapan and BN. Only a minority of PH/PR seats were won with less than 50% of the votes 

<br>


```{r three-cornered}
ballots |> 
  # Beginning of the collapse of BN and the opening up of 
  # the political space
  filter(date > "2013-01-01" & votes_perc >= 5) |> 
  mutate(count = 1) |> 
  group_by(seat, date, federal) |> 
  summarise(candidates = sum(count), .groups = "drop") |> 
  mutate(multicandidate = ifelse(
    candidates > 2, 1, 0
  )) |> 
  left_join(
    ballots |> 
      filter(date > "2013-01-01" & 
               result %in% c("won", "won_uncontested")) |> 
      select(date, seat, winner = coalition), 
    by = c("date", "seat")
  ) |> 
  filter(candidates != 1) |> 
  mutate(count = 1) |> 
  ggplot(aes(x = candidates, y = count)) + 
  geom_col(width = 1, 
                 aes(fill = winner)) +
  facet_wrap(~ federal, scales = "free") + 
  scale_fill_viridis_d(option = "magma") + 
  scale_x_continuous(breaks = seq(0, 7, 1)) + 
  theme(strip.background = element_rect(fill = "grey20")) + 
  labs(title = "Three-cornered fights are usually won by PH or BN", 
       subtitle = "Only electoral races post UMNO collapse (2018); only counts candidates which won more than 5% of the vote share", 
       y = "Number of electoral races", 
       n = "Number of candidates", 
       fill = "Winner")
  
  

```

<br>

However, the margin by which PH/PR has won has been trending downwards. And its footprint is narrowing, especially with PKR focusing on [safe seats](https://www.freemalaysiatoday.com/category/nation/2026/04/16/avoiding-weak-seats-a-realistic-strategy-says-analyst).

<br>



```{r}
ballots |> 
  filter(coalition %in% c("PH", "PR")) |> 
  mutate(year = year(date)) |> 
  ggplot(aes(x = str_sub(year, 3L), y = votes_perc / 100)) + 
  geom_jitter(aes(colour = result)) + 
  scale_colour_viridis_d(option = "plasma") + 
  scale_y_continuous(labels = percent, 
                     breaks = seq(0, 1, .1)) +
  facet_wrap(~ federal) + 
  theme(strip.background = element_rect(fill = "grey20")) + 
  labs(x = "Year", 
       y = "PH/PR vote share", 
       title = "Electoral performance of PH by vote share")

```

<br>


[James Chin](https://www.commonwealthroundtable.co.uk/commonwealth/eurasia/malaysia/research-article-anwars-long-walk-to-power-the-2022-malaysian-general-elections/) at the Commonwealth Roundtable argues that there is no longer any middle ground left in Malaysian politics, with the DAP on one end and PAS on the other. This erosioon of the middle ground will have a profound impact on Pakatan Harapan's election chances, especially as it refuses to engage in the Islamisation race or put forward a vision that can win over Malay youth. 

<br><br><br>

# Data 

The data for this report comes primarily from  [Thevesh](https://github.com/Thevesh/paper-meco-results/tree/main), who compiled the *Malaysian Election Corpus*, containing federal and state-level election results since 1955. 

Additional data, including the digitised census datasets, come from [JJean95](https://www.kaggle.com/datasets/jjean95/malaysia-general-election-datasets?select=census_dun.csv).

In this analysis, we have used constituency populations as opposed to the number of voters because MPs represent all peoples within their constituencies, not just electors. 

<br><br><br>


# References

::: {#refs}

Architects of Diverity (2023). *State of Discrimination Survey 2023*. Retrieved from https://www.aodmalaysia.org/sods. 

Azzubair, Kartini & Nazri Muslim (2026). A Systematic Literature Review of Malaysia's Coalition Politics, 2021-2025. *Frontiers in Political Science, Elections and Representation section*, 8-2026. https://doi.org/10.3389/fpos.2026.1721966

Chin, J. (2023) Anwar’s long walk to power: the 2022 Malaysian general
elections, The Round Table, 112:1, 1-13, DOI: 10.1080/00358533.2023.2165303

JJean95 (2022). *Malaysia General Election Results (GE12-GE15)*. https://www.kaggle.com/datasets/jjean95/malaysia-general-election-datasets?select=census_dun.csv

Deivasagayam, A.D. (2025). Explaining UMNO's Downfall Post GE14 and GE15: The Strengthened Convergence Between Two Cleavage Systems. *Malaysian Journal of Social Sciences and Humanities*, 10(2). https://doi.org/10.47405/mjssh.v10i2.3248

Nora M. (2026, 16 April). Avoiding Weak Seats a Realistic Strategy, Says Analyst. *Free Malaysia Today*. https://www.freemalaysiatoday.com/category/nation/2026/04/16/avoiding-weak-seats-a-realistic-strategy-says-analyst

Pusat KOMAS (2024). *Malaysia Racial Discrimination Report: a Decade in review (2015-2024)*. Retrieved from https://komas.org/download/Malaysia-Racism-Report_A-Decade-In-Review-2015-2024.pdf

Syaza S. (2024). Why Young Malay Voters in Malaysia are "Turning Green". *Trends in Southeast Asia, ISEAS -- Yusof Ishak Institute*, 12. Retrieved from https://www.iseas.edu.sg/wp-content/uploads/2024/04/TRS12_24.pdf

Thevesh T. (2026). *Malaysian Election Corpus: Federal and State-Level Election Results since 1955*. https://github.com/Thevesh/paper-meco-results/tree/main


:::

