---
title: "Malaysian Elections"
subtitle: "Part one: Ethnicity"
author: "Sean Ng"
organization: "AIMdata"
date-modified: "11 January 2026"
execute: 
  echo: false
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 9)


library(tidyverse)
library(here)
library(janitor)
library(scales)
library(tidytext)
library(widyr)
library(ggraph)
library(patchwork)
library(kableExtra)
library(viridis)
library(stringr)
library(sf)
library(ggmagnify)

`%out%` <- Negate(`%in%`)
options(scipen = 100)
theme_set(theme_light())
range_wna <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

last_comma <- function(x, sep = ","){
  x <- sapply(strsplit(x, sep),"[")
  trimws(x[[length(x)]])
}
```




```{r data}

candidates <- rbind(
  read_csv("./data/candidates_ge12.csv") |>
    mutate(election = "ge12", new_mp = NA_real_), 
  read_csv("./data/candidates_ge13.csv") |>
    mutate(election = "ge13", new_mp = NA_real_), 
  read_csv("./data/candidates_ge14.csv") |>
    mutate(election = "ge14", new_mp = NA_real_), 
  read_csv("./data/candidates_ge15.csv") |>
    mutate(election = "ge15") |>
    select(-name) |>
    rename(name = name_display)) 

results <- rbind(
  read_csv("./data/results_parlimen_ge12.csv"),
  read_csv("./data/results_parlimen_ge13.csv"),
  read_csv("./data/results_parlimen_ge14.csv"),
  read_csv("./data/results_parlimen_ge15.csv")
)

census_dun <- read_csv("./data/census_dun.csv") |> 
  mutate(population_bumi = ethnicity_proportion_bumi * population_total / 100, 
         population_chinese = ethnicity_proportion_chinese * population_total/ 100, 
         population_indian = ethnicity_proportion_indian * population_total / 100, 
         population_other = ethnicity_proportion_other * population_total / 100)

voters_ge15 <- read_csv("./data/voters_ge15.csv")

candidates <- candidates |> 
  left_join(
    candidates |> 
      group_by(parlimen, election) |> 
      summarise(total_votes = sum(votes)), 
    by = c("parlimen", "election")
  ) |> 
  mutate(vote_share = votes / total_votes) |> 
  ungroup() |> 
  # Result in errors where both candidates won
  mutate(result = 
           case_when(
             votes == 15495 & election == "ge13" & parlimen == "P.170 Tuaran" ~ 
               0, 
             votes == 12155 & election == "ge13" & parlimen == "P.177 Beaufort" ~ 
               0, 
             votes == 11900 & election == "ge13" & parlimen == "P.180 Keningau" ~ 
               0, 
             TRUE ~ result), 
         result_dec = 
           case_when(
             votes == 15495 & election == "ge13" & parlimen == "P.170 Tuaran" ~ 
               "lost", 
             votes == 12155 & election == "ge13" & parlimen == "P.177 Beaufort" ~ 
               "lost", 
             votes == 11900 & election == "ge13" & parlimen == "P.180 Keningau" ~ 
               "lost", 
             TRUE ~ result_desc))

gini <- read_csv("./data/hh_inequality_parlimen.csv") |> 
  filter(date == "2024-01-01")


sf::sf_use_s2(FALSE)

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))

federal_territories <- tribble(
  ~parlimen, ~population_total, ~pc_bumi, ~pc_chinese, ~pc_indian, ~pc_other, ~sex_male, ~sex_female, ~population_density, ~income_avg, ~expenditure_avg, ~poverty_incidence, ~area_km2,
  "P.114 Kepong", 106199, .202, .729, .063, .006, .588, .412, 8850, 13087, 6500, .003, 12,
  "P.115 Batu", 219132, .566, .301, .129, .004, .601, .399, 10957, 11019, 5817, 0, 20,
  "P.116 Wangsa Maju", 215600, .617, .306, .07, .008, .537, .463, 13475, 11153, 5768, 0, 16,
  "P.117 Segambut", 253715, .324, .558, .107, .012, .512, .479, 4975, 20521, 9845, 0, 51,
  "P.118 Setiawangsa", 147095, .697, .214, .085, .004, .519, .481, 9193, 12281, 6534, .003, 16, 
  "P.119 Titiwangsa", 122096, .677, .233, .076, .014, .519, .481, 8140, 11921, 6823, .005, 15,
  "P.120 Bukit Bintang", 120529, .271, .565, .155, .009, .565, .435, 5739, 13677, 7401, 0, 21, 
  "P.121 Lembah Pantai", 148094, .591, .262, .138, .009, .509, .491, 7405, 13081, 6186, 0, 20,
  "P.122 Seputeh", 322511, .311, .573, .112, .004, .510, .490, 10404, 13489, 7764, .005, 31,
  "P.123 Cheras", 135823, .322, .596, .076, .007, .527, .473, 8489, 12031, 5979, .002, 16,
  "P.124 Bandar Tun Razak", 191318, .679, .234, .083, .004, .509, .491, 7653, 11266, 6002, 0, 25,
  "P.166 Labuan", 95120, .862, .117, .011, .01, .514, .486, 1034, 8319, 4097, .031, 96,
  "P.125 Putrajaya", 109202, .979, .012, .006, .003, .509, .491, 2229, 12840, 7980, .004, 49
  
)

census_fed <- census_dun |> 
  mutate(population_poor = poverty_incidence * population_total / 100) |> 
  group_by(parlimen) |> 
  summarise_at(
    c("population_total", 
      "population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ sum(.x, na.rm = TRUE)
  ) |> 
  mutate_at(
    c("population_bumi", 
      "population_chinese", 
      "population_indian", 
      "population_other", 
      "population_poor", 
      "sex_male", 
      "sex_female"), 
    ~ .x / population_total
  ) |>  
  rename(
    pc_bumi = population_bumi, 
    pc_chinese = population_chinese, 
    pc_indian = population_indian, 
    pc_other = population_other, 
    poverty_incidence = population_poor) |> 
  left_join(
    census_dun |>
      mutate(
        income = income_avg * household_total,
        expenditure = expenditure_avg * household_total) |>
      group_by(parlimen) |>
      summarise(
        household_total = sum(household_total),
        income = sum(income),
        expenditure = sum(expenditure)) |>
      mutate(
        income_avg = income / household_total,
        expenditure_avg = expenditure / household_total) |>
      select(parlimen,
             income_avg, 
             expenditure_avg), 
    by = "parlimen") |> 
  left_join(
    parlimen_geo |> 
      st_drop_geometry() |> 
      select(parlimen, area_km2), 
    by = "parlimen"
  ) |> 
  mutate(population_density = population_total / area_km2) |> 
  rbind(federal_territories) |> 
  left_join(
    census_dun |> 
      distinct(parlimen, state), 
    by = "parlimen"
  ) |>
  left_join(
    gini |> 
      select(parlimen, gini), 
    by = "parlimen"
  ) |> 
  mutate(state = case_when(
    is.na(state) & str_detect(parlimen, "Labuan") ~ "Labuan", 
    is.na(state) & str_detect(parlimen, "Putrajaya") ~ "Putrajaya", 
    is.na(state) ~ "Kuala Lumpur", 
    TRUE ~ state
  )) |> 
  mutate(east_malaysia = ifelse(str_detect(state, "Sabah|Sarawak|Labuan"), 
                                1, 0))

ballots <- read_csv("./data/consol_ballots.csv") |> 
  mutate(
    age = ifelse(age == -1, NA_real_, age)
  ) |> 
  mutate(federal = ifelse(
    str_detect(seat, "P.0|P.1|P.2"), 
    "Federal", 
    "State"))

stats <- read_csv("./data/consol_stats.csv")

parlimen_geo <- read_sf("./data/parlimen.geojson") |> 
  st_make_valid() |> 
  mutate(area_km2 = st_area(geometry) / 1000000, 
         area_km2 = as.numeric(area_km2))
```


# Racialisation of elections


```{r}
multiracial |> 
  mutate(count = 1) |> 
  filter(federal == "Federal") |> 
  summarise_at(
    vars(malay_only, chinese_only, indian_only, 
         sabah_only, sarawak_only, other_only, 
         multiracial, count), 
    ~ sum(.x, na.rm = TRUE)
  ) |> 
  pivot_longer(!count) |> 
  mutate(pc = value / count) |> 
  mutate(name = str_to_title(str_replace_all(name, "_", " "))) |> 
  ggplot(aes(x = pc, y = fct_reorder(name, pc))) + 
  geom_col(fill = "#e9d8a6") + 
  geom_text(aes(label = comma(value)), 
            hjust = "inward") +
  scale_x_continuous(labels = percent) + 
  labs(
    x = "% of federal elections", 
    y = "", 
    title = "Ethnicity of candidates in federal elections"
  )
  
  
```

As have already seen, most constituencies in Malaysia are multi-ethnic in nature: `r census_fed |> filter(pc_bumi > .8) |> nrow()` out of the `r census_fed |> nrow()` federal constituencies have at least 20% of their populations being racial minorities. 

We'll set the threshold for being included in analysis at 3% of votes 

```{r}
multiracial_races <- ballots |> 
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(count = 1) |> 
  # I think this filter is fine
  filter(rank <= 3) |> 
  group_by(state, seat, date, federal, election) |> 
  summarise(malay_candidates = sum(count[ethnicity == "Malay"]), 
            chinese_candidates = sum(count[ethnicity == "Chinese"]), 
            indian_candidates = sum(count[ethnicity == "Indian"]), 
            sabah_candidates = sum(count[ethnicity == "Bumi Sabah"]), 
            sarawak_candidates = sum(count[ethnicity == "Bumi Sarawak"]), 
            other_candidates = sum(count[ethnicity == "Other"]), 
            .groups = "drop") |> 
  mutate(total_candidates = malay_candidates + chinese_candidates + indian_candidates + 
           sabah_candidates + sarawak_candidates + other_candidates) |> 
  pivot_longer(cols = malay_candidates:other_candidates, 
               names_to = "candidate_ethnicity", 
               values_to = "value") |> 
  mutate(candidate_ethnicity = str_remove_all(candidate_ethnicity, "_candidates")) |> 
  filter(value != 0) |>
  mutate(multiracial = ifelse(value != total_candidates, 1, 0)) |> 
  mutate(combination = ifelse(
    multiracial == 0, paste0(candidate_ethnicity, "_only"), "multiethnic"
  )) |> 
  distinct(state, seat, date, federal, election, multiracial, combination)
```

```{r}
ballots |>
  left_join(
    multiracial_races, 
    by = c("state", "seat", "date", "federal", "election")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1) |>
  filter(result %in% c("won", "won_uncontested")) |> 
  group_by(
    decade, federal, combination
  ) |> 
  summarise(count = sum(count), .groups = "drop") |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count), 
         combination = case_when(
           combination == "sabah_only" ~ "bumi_sabah_only", 
           combination == "sarawak_only" ~ "bumi_sarawak_only", 
           TRUE ~ combination), 
         combination = str_to_title(str_replace_all(combination, "_", " "))
         ) |> 
  ungroup() |>
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, y = pc, fill = fct_relevel(combination, 
                                   c("Bumi Sabah Only", 
                                     "Bumi Sarawak Only", 
                                     "Chinese Only", 
                                     "Indian Only", 
                                     "Malay Only", 
                                     "Other Only", 
                                     "Multiethnic")))) + 
  geom_col() + 
  facet_wrap(~ federal) + 
  scale_y_continuous(labels = percent) + 
  scale_fill_viridis_d() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(fill = "Combination",
       title = "Multiethnic electoral races were least common during the 90s and 00s", 
       x = "", 
       y = "") + 
  theme(strip.background = element_rect(fill = "grey20"))
```

Multiethnic were at their height in the 1950s. 

```{r}

```

```{r}
parlimen_geo |> glimpse()
```



```{r}
parlimen_geo |> 
  left_join(
    multiracial_races |>
      filter(federal == "Federal") |>
      select(seat, multiracial, combination) |>
      group_by(seat) |>
      slice_max(order_by = multiracial) |>
      ungroup() |>
      distinct(seat, multiracial, combination), 
    by = c("parlimen" = "seat") 
  ) |> 
  ggplot() + 
  geom_sf(aes(fill = multiracial), 
          linewidth = .1) + 
  # Only one inset can work at one time
  geom_magnify(
    from = c(101.223422,  101.953542, 2.805661, 3.383667), 
    to = c(105, 108, 3.5, 5.9), 
    shadow = FALSE,
    proj = "single"
  ) + 
  theme_void()

```





```{r}
ballots |> 
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(count = 1) |> 
  # I think this filter is fine
  filter(rank <= 3) |> 
  group_by(state, seat, date, federal, election) |> 
  summarise(malay_candidates = sum(count[ethnicity == "Malay"]), 
            chinese_candidates = sum(count[ethnicity == "Chinese"]), 
            indian_candidates = sum(count[ethnicity == "Indian"]), 
            sabah_candidates = sum(count[ethnicity == "Bumi Sabah"]), 
            sarawak_candidates = sum(count[ethnicity == "Bumi Sarawak"]), 
            other_candidates = sum(count[ethnicity == "Other"]), 
            .groups = "drop") |> 
  mutate(total_candidates = malay_candidates + chinese_candidates + indian_candidates + 
           sabah_candidates + sarawak_candidates + other_candidates) |> 
  pivot_longer(cols = malay_candidates:other_candidates, 
               names_to = "candidate_ethnicity", 
               values_to = "value") |> 
  mutate(candidate_ethnicity = str_remove_all(candidate_ethnicity, "_candidates")) |> 
  filter(value != 0) |>
  mutate(multiracial = ifelse(value != total_candidates, 1, 0)) |> 
  mutate(combination = ifelse(
    multiracial == 0, paste0(candidate_ethnicity, "_only"), "multiethnic"
  ))
```


```{r}
multiracial <- ballots |> 
  mutate(ethnicity = ifelse(ethnicity == "Orang Asli", "Other", ethnicity)) |> 
  mutate(count = 1) |> 
  # I think this filter is fine
  filter(rank <= 3) |> 
  group_by(state, seat, date, federal, election) |> 
  summarise(malay_candidates = sum(count[ethnicity == "Malay"]), 
            chinese_candidates = sum(count[ethnicity == "Chinese"]), 
            indian_candidates = sum(count[ethnicity == "Indian"]), 
            sabah_candidates = sum(count[ethnicity == "Bumi Sabah"]), 
            sarawak_candidates = sum(count[ethnicity == "Bumi Sarawak"]), 
            other_candidates = sum(count[ethnicity == "Other"]), 
            .groups = "drop") |> 
  mutate(malay_chinese = ifelse(malay_candidates > 0 & chinese_candidates > 0, 1, 0), 
         malay_indian = ifelse(malay_candidates > 0 & indian_candidates > 0, 1, 0), 
         chinese_indian = ifelse(chinese_candidates > 0 & indian_candidates > 0, 1, 0), 
         malay_sabah = ifelse(malay_candidates > 0 & sabah_candidates > 0, 1, 0), 
         malay_sarawak = ifelse(malay_candidates > 0 & sarawak_candidates > 0, 1, 0), 
         chinese_sabah = ifelse(chinese_candidates > 0 & sabah_candidates > 0, 1, 0), 
         chinese_sarawak = ifelse(chinese_candidates > 0  & sarawak_candidates > 0, 1, 0), 
         indian_sabah = ifelse(indian_candidates > 0 & sabah_candidates > 0, 1, 0), 
         indian_sarawak = ifelse(indian_candidates > 0 & sarawak_candidates > 0, 1, 0), 
         malay_other = ifelse(malay_candidates > 0 & other_candidates > 0, 1, 0), 
         chinese_other = ifelse(chinese_candidates > 0 & other_candidates > 0, 1, 0), 
         indian_other = ifelse(indian_candidates > 0 & other_candidates > 0, 1, 0), 
         sabah_other = ifelse(sabah_candidates > 0 & other_candidates > 0, 1, 0), 
         sarawak_other = ifelse(sarawak_candidates > 0 & other_candidates > 0, 1, 0), 
         
         malay_chinese_indian = ifelse(malay_candidates > 0 & chinese_candidates > 0 & 
                                         indian_candidates > 0, 1, 0), 
         malay_chinese_other = ifelse(malay_candidates > 0 & chinese_candidates > 0 & 
                                        other_candidates > 0, 1, 0), 
         malay_chinese_sabah = ifelse(malay_candidates > 0 & chinese_candidates > 0 & 
                                        sabah_candidates > 0, 1, 0), 
         malay_chinese_sarawak = ifelse(malay_candidates > 0 & chinese_candidates > 0 & 
                                          sarawak_candidates > 0, 1, 0), 
         chinese_sabah_sarawak = ifelse(chinese_candidates > 0 & sabah_candidates > 0 & 
                                          sarawak_candidates > 0, 1, 0), 
         sabah_sarawak_other = ifelse(sabah_candidates > 0 & sarawak_candidates > 0 & 
                                        other_candidates > 0, 1, 0),
         
         malay_only = ifelse(malay_candidates > 0 & chinese_candidates == 0 & 
                               indian_candidates == 0 & sabah_candidates == 0 & 
                               sarawak_candidates == 0 & other_candidates == 0, 
                             1, 0), 
         chinese_only = ifelse(malay_candidates == 0 & chinese_candidates > 0 & 
                               indian_candidates == 0 & sabah_candidates == 0 & 
                               sarawak_candidates == 0 & other_candidates == 0, 
                             1, 0),
         indian_only = ifelse(malay_candidates == 0 & chinese_candidates == 0 & 
                               indian_candidates > 0 & sabah_candidates == 0 & 
                               sarawak_candidates == 0 & other_candidates == 0, 
                             1, 0),
         sabah_only = ifelse(malay_candidates == 0 & chinese_candidates == 0 & 
                               indian_candidates == 0 & sabah_candidates > 0 & 
                               sarawak_candidates == 0 & other_candidates == 0, 
                             1, 0),
         sarawak_only = ifelse(malay_candidates == 0 & chinese_candidates == 0 & 
                               indian_candidates == 0 & sabah_candidates == 0 & 
                               sarawak_candidates > 0 & other_candidates == 0, 
                             1, 0),
         other_only = ifelse(malay_candidates == 0 & chinese_candidates == 0 & 
                               indian_candidates == 0 & sabah_candidates == 0 & 
                               sarawak_candidates == 0 & other_candidates > 0, 
                             1, 0)
         ) |> 
  select(state, seat, date, federal, election, malay_chinese:other_only) |> 
  mutate(multiracial = ifelse(malay_only == 1 | chinese_only == 1 | indian_only == 1 |
                                sabah_only == 1 | sarawak_only == 1 | other_only == 1,
                              0, 1))  
  



```




```{r}
multiracial |> 
  filter(multiracial == 1) |> 
  left_join(
    ballots |> 
      filter(result %in% c("won", "won_uncontested")) |> 
      select(seat, date, winner_ethnicity = ethnicity), 
    by = c("seat", "date")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |> 
  select(-c(malay_only:other_only)) |> 
  pivot_longer(
    cols = malay_chinese:sabah_sarawak_other, 
    names_to = "combination", 
    values_to = "value"
  ) |> 
  filter(value > 0) |> 
  mutate(
    combination = ifelse(
      combination %in% c(
        "malay_chinese_other", 
        "malay_sarawak", 
        "indian_other", 
        "sabah_other", 
        "sarawak_other", 
        "malay_chinese_other", 
        "malay_chinese_sabah", 
        "malay_chinese_sarawak"
      ), 
      "other_combination", 
      combination
    )) |> 
  group_by(combination) |> 
  mutate(sort = sum(value)) |> 
  ungroup() |> 
  mutate(combination = str_replace_all(combination, "_", "/"), 
         combination = str_to_title(combination)) |> 
  ggplot(aes(
    x = value, 
    y = fct_reorder(combination, sort), 
    fill = winner_ethnicity
  )) + 
  geom_col() +
  scale_fill_viridis_d() +
  labs(title = "Multiethnic electoral races and their results",
       x = "Number of electoral races", 
       y = "", 
       fill = "Winner ethnicity") +
  facet_wrap(~ federal, scales = "free_x") + 
  theme(strip.background = element_rect(fill = "grey20"))
```



```{r fig.height=7}
multiracial |> 
  filter(multiracial == 1) |> 
  left_join(
    ballots |> 
      filter(result %in% c("won", "won_uncontested")) |> 
      select(seat, date, winner_ethnicity = ethnicity), 
    by = c("seat", "date")
  ) |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |> 
  select(-c(malay_only:other_only)) |> 
  pivot_longer(
    cols = malay_chinese:sabah_sarawak_other, 
    names_to = "combination", 
    values_to = "value"
  ) |> 
  mutate(winner_ethnicity = ifelse(
    winner_ethnicity == "Orang Asli", "Other", winner_ethnicity
  )) |> 
  ggplot(aes(x = decade, 
             y = value, 
             fill = winner_ethnicity)) + 
  geom_col() + 
  scale_fill_viridis_d() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) + 
  facet_wrap(~ federal, scales = "free_y") + 
  labs(
    title = "Results of multiethnic electoral races"
  )

```

```{r}
multiracial |> 
  mutate(year = year(date), 
         decade = year - year %% 10) |>
  group_by(decade, federal) |> 
  summarise(n = n(), 
            multiethnic = sum(multiracial), 
            .groups = "drop")  |> 
  mutate(single_race = n - multiethnic,
         multiethnic_pc = multiethnic/ n, 
         single_race_pc = single_race/ n) |> 
  select(decade, federal, n, multiethnic_pc, single_race_pc) |> 
  pivot_longer(cols = multiethnic_pc:single_race_pc) |>
  ggplot(aes(x = decade, y = value, fill = name)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  facet_wrap(~ federal, scales = "free_y") +
  scale_fill_viridis_d(option = "cividis", begin = .2, end = .9)
  
  glimpse()
```

Instead of east malaysia, maybe by multiethnic fight?

Let's take a look at Chinese candidates, the largest minority group

non-Malay representation?

```{r}
ballots |> glimpse()
```

```{r}
ballots |> 
  mutate(votes_perc = ifelse(result == "won_uncontested", 100, votes_perc)) |> 
  filter(votes_perc > 3 & ethnicity == "Chinese") |> 
  mutate(count = 1) |> 
  group_by(party, federal) |> 
  summarise(count = sum(count)) |> 
  arrange(desc(count))
```

```{r}
ballots |> 
  mutate(count = 1) |> 
  
```




```{r}
candidates |> 
  mutate(east_malaysia = 
           ifelse(str_detect(state, 
                             "Sabah|Sarawak|Labuan"), 
                  "Yes", "No")) |> 
  group_by(parlimen, east_malaysia, election) |> 
  summarise(
    total_votes = sum(votes), 
    winning_votes = sum(votes[result == 1]), 
    losing_votes = sum(votes[result != 1]), 
    .groups = "drop"
  ) |> 
  mutate(margin = (winning_votes - losing_votes) / total_votes) |> 
  ggplot(aes(x = margin, fill = fct_rev(east_malaysia), group = east_malaysia)) + 
  geom_histogram() + 
  scale_x_continuous(labels = percent) + 
  scale_fill_manual(values = c("#90e0ef", "#03045e")) + 
  labs(x = "Margin of victory", 
       y = "Number of electoral races", 
       title = "Most electoral races in Malaysia are competitive", 
       subtitle = "Margins below 0% are three-cornerned fights", 
       fill = "East Malaysia")

```


Map of where multiracial races are

```{r}
multiracial |>
  filter(federal == "Federal") |> 
  mutate(year = year(date),
         decade = year - year %% 10,
         count = 1) |> 
  group_by(decade) |> 
  summarise(count = sum(count), 
            multiracial = sum(multiracial), 
            .groups = "drop") |> 
  mutate(pc = multiracial / count) |> 
  ggplot(aes(x = decade, y = pc)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) + 
  scale_y_continuous(labels = percent) + 
  labs(x = "", 
       y = "% of multiracial election races", 
       title = "% of Federal election races")
  
```

```{r}
multiracial |> glimpse()
```
```{r}
ballots |> count(election)

multiracial |> glimpse()
  
```


```{r fig.height=8}
census_fed |> 
  left_join(
    multiracial |> 
      filter(election %in% c("GE-12", "GE-13", "GE-14", "GE-15")) |> 
      mutate(type = case_when(
        multiracial == 1 ~ "Multiethnic", 
        malay_only == 1 ~ "Malay only", 
        chinese_only == 1 ~ "Chinese only", 
        indian_only == 1 ~ "Indian only", 
        sabah_only == 1 ~ "Bumi Sabah only", 
        sarawak_only == 1 ~ "Bumi Sarawak only"
      )) |> 
      select(seat, type, election), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_bumi = population_total * pc_bumi) |> 
  ggplot(aes(x = population_total, 
             y = pc_bumi)) + 
  geom_point(aes(colour = type, 
                 size = population_bumi)) +
  scale_x_log10(labels = comma) + 
  scale_y_continuous(labels = percent) +
  scale_colour_viridis_d() +
  scale_size_continuous(labels = comma) +
  labs(x = "Population of constituency", 
       y = "% of population who are Bumiputera", 
       colour = "Ethnicity of\ncandidates", 
       size = "2020 Bumiputera\npopulation") +
  guides(colour = guide_legend(
    override.aes = list(size = 2)
  )) + 
  facet_wrap(~ election)
```


```{r}
census_fed |> 
  left_join(
    multiracial |> 
      filter(election == "GE-15") |> 
      mutate(type = case_when(
        multiracial == 1 ~ "Multiethnic", 
        malay_only == 1 ~ "Malay only", 
        chinese_only == 1 ~ "Chinese only", 
        indian_only == 1 ~ "Indian only", 
        sabah_only == 1 ~ "Bumi Sabah only", 
        sarawak_only == 1 ~ "Bumi Sarawak only"
      )) |> 
      select(seat, type), 
    by = c("parlimen" = "seat")
  ) |> 
  mutate(population_bumi = population_total * pc_bumi) |> 
  ggplot(aes(x = population_density, 
             y = pc_bumi)) + 
  geom_point(aes(colour = type, 
                 size = population_total)) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = percent) +
  scale_colour_viridis_d() +
  scale_size_continuous(labels = comma) +
  labs(x = "Population density", 
       y = "% of population who are Bumiputera", 
       colour = "Ethnicity of\ncandidates", 
       size = "Total\npopulation") +
  guides(colour = guide_legend(
    override.aes = list(size = 3)
  ))
```

<br>

No Malay candidates ran where the population of non-Malays were 

Maybe put a red circle around the constituencies which were particularly competitive

```{r}
ballots |> 
  filter(rank == 3 & votes_perc > .03 & federal == "Federal") 

ballots |> filter(votes_perc > .1 & federal == "Federal") |> 
  group_by(date, seat, ethnicity) |> 
  summarise(votes = sum(votes))
```



<br>

```{r}

census_fed |> 
  mutate_at(
    vars("pc_bumi", "pc_chinese", "pc_indian", "pc_other", 
         "poverty_incidence", "sex_male", "sex_female"),
    ~ .x * population_total
  ) |> 
  rename_all(
    .funs = ~ gsub("pc_", "", .)) |> 
  pivot_longer(
    cols = bumi:other, 
    names_to = "ethnicity", 
    values_to = "value"
  ) |> 
  mutate(ethnicity = str_to_title(ethnicity)) |> 
  select(parlimen, 
         ethnicity, 
         value, 
         population_total) |> 
  ggplot(aes(y = value, fill = ethnicity, 
             x = fct_reorder(str_sub(parlimen, start = 1L, end = 5L), 
                             population_total))) + 
  geom_col(width = 1) + 
  scale_fill_viridis_d(option = "plasma") + 
  theme(axis.text.x = element_text(size = .5, angle = 90)) +
  scale_y_continuous(labels = number_format(scale = .001, suffix = "k"), 
                     breaks = seq(0, 700000, 50000)) + 
  labs(x = "Parliamentary seat", 
       y = "Constituency population", 
       title = "Massive parliamentary misapportionment", 
       subtitle = "A vote in the smallest constituency (Kanowit) is is worth 27.8 votes in the largest constituency (Bangi)", 
       fill = "")

```

```{r}
ballots |> 
  filter(state %in% c("Sabah", "Sarawak")) |> 
  mutate(ethnicity = ifelse(
    ethnicity == "Orang Asli", "Other", ethnicity
  )) |> 
  mutate(year = year(date), 
         decade = year - year %% 10, 
         count = 1) |> 
  group_by(state, decade, ethnicity, federal) |> 
  summarise(count = sum(count)) |> 
  group_by(decade, federal) |> 
  mutate(total = sum(count)) |> 
  ungroup() |> 
  mutate(pc = count / total) |> 
  ggplot(aes(x = decade, 
             y = pc, 
             fill = ethnicity)) + 
  geom_col(position = "stack") + 
  facet_wrap(~ federal + state) + 
  scale_y_continuous(labels = percent) + 
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  scale_fill_viridis_d() + 
  theme(axis.text.x = element_text(angle = 30, vjust = .5), 
        strip.background = element_rect(fill = "black")) + 
  labs(title = "Ethnicity of election candidates", 
       y = "% of candidates", 
       x = "") 
```




